---
title: "Tables in Pharma"
author: "R Consortium Workstream Tables for Regulatory Submission"
date: "6/3/2021"
output:
  html_document: default
  pdf_document: default
subtitle: Cell Values Derivation and Simple Tabulation
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Tables are a major component for analyzing clinical trials data and to communicate the results to the study teams, in
publications, and to the health authorities. Historically most tables for clinical trials submissions have been created
with SAS but there is an ongoing industry effort to switch the programming language of choice to R.

The R Consortium has a number of working streams to coordinate the efforts among Pharma companies and health authorities
to establish the requirements for an R based submission (the analytics part). The aim is to spawn collaborations on open
source solutions that will settle in as the industry standard.

This paper on table creation is the first paper from the  *R Consortium Tables for Regulatory Submission (RTRS)* working
group. The aim is to publish a series of papers on table creation to analyze clinical trials data with R. The series of
papers will be:

1. Data Derivation for Tabulation & Table Data Structure
   - (`dplyr` & `tidyr`, `rtables`, `Tplyr`, `tables`)
   - Analytics & matching results (rounding rules)
   - data structure & accessors and modifiers
1. Table creation Process -- Industry
   - stages
1. Formatting & Rendering
   - cell format
   - (tables, titles and footnotes) & Rendering (html, pdf, txt, rtf, xml, markdown & rmarkdown, word, power point, tgen etc.)
      - including `pharmaRTF`
   - pagination
   - titles, footnotes, cell references
3. Listings

1. Creating Tables at Scale (Workflow, pipelining)
   - table templates (standards)
   - cross checks
   - hybrid SAS and R projects
   - saving table cell values
1. clinicaltrials.gov submission
1. Future Recommendations
   - Interactive Tables and drill down capability
      - shiny
      - e.g. patient profile table, listings, graphs
   - HTML website with tables as reviewing aid
      - add more markup/formatting
      - search/organizing

1. Table Catalog??
1. Transforming tables??

In this paper we discuss key data derivation concepts behind tabulation.


# Table R packages Overview

R's table creation functionality has recently seen a lot of contribution via specialized R packages. In this section we
provide an overview of packages what this working group has found to be useful in analyzing clinical trials data.

Open source projects in alphabetical order (remember, this is not a sales pitch, please be as objective as possible in describing the functionality of the package):

- `broom`: useful to create tables of statistical model fits
- `flextable`:
- `gt`:
- `huxtable`:
- `mmtable2`:
- `rtables`: defines a pipe-able grammar for declaring complex table structure layouts via nested splitting - in
row- and column-space - and the specification of analysis functions which will calculate cell values within that
structure. These pre-data table layouts are then applied to data to build the table, during which all necessary data
grouping implied by the row and column splits, cell value calculation, and tabulation occurs automatically. Additional
table features such as titles, footers, and referential footnotes are supported. ASCII, HTML, and PDF are supported as
output formats for rendering tables.
- `tableone`: 
- `tables`: 
- `tplyr`:

`pharmaRTF`, `xtable`, `r2rtf`, `officer`, `pander`, `ttt` fit in to the discussion of the 2nd paper.

Note that `DT`, `reactable` are not listed here as it is used to create interactive tables only. 


Other R packages?:

# What is a Table

https://uwspace.uwaterloo.ca/bitstream/handle/10012/10962/WANG_XINXIN_.pdf?sequence=1&isAllowed=y

The tables of interest for this paper are those who represent summarizations of data.

Computing a Table:

* Derive Cell Values from Data
  - this includes splitting data to subsets for each cell
  - calculate cell values from data subsets
* Tabulation: mapping to table structure
  - row and column labeling
  - i.e. arrange cell values in a tabular form
  - titles, footnotes
  - cell, row, column references
* Rendering & Formatting:
   - to a device (display, HTML, PDF,  txt, ...)
   - colors

Presentational Form Table:

*A table is an arrangement of data in rows and columns, or possibly in a more complex structure.* (Wikipedia)

* Cells
* Columns
* Rows
* ...

# Other Important Requirements For Pharma

* Pagination
* regulated (validation)

# Data Used

```{r, message=FALSE}
adsl <- rtables::ex_adsl
adae <- rtables::ex_adae
```

# Cell Content Derivation

For this paper we look at cell content derivation. The different table packages (except for `broom`) can be grouped into
frameworks that require the cell content to be per-calculated and available in a data frame and frameworks that come
with specialized cell derivation methods.

The table packages that rely on external ways of deriving the cell contents include `gt`, `flextable`, `huxtable`,
`xtable`. We found that `dplyr` and `tidyr` are well suited for that task and use those two packages for the further
cell derivation discussion. In the subsequent papers we will discuss the tabulation, rendering, and other relevant
features of those table packages.

Packages that provide specialized cell value derivation tooling include: base R, `tables`, `Tplyr`, `rtables`.

We now provide ... examples that discuss different requirements of cell content derivation and basic tabulation in form
of row and column arrangement.


## Cross Tabulation

```{r}
library(dplyr)
library(tidyr)
library(tables)
library(Tplyr)
library(rtables)
```

Calculate frequencies

```{r}
# base R `table` function
table(adsl$COUNTRY, adsl$ARM)
```


```{r}
# base R `xtabs`
xtabs(~ COUNTRY + ARM, data = adsl)
```


```{r}
# `dplyr` and `tidyr`
adsl %>%
  group_by(ARM, COUNTRY) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  pivot_wider(names_from = ARM, values_from = n)
```


```{r}
# `rtables`
lyt <- basic_table() %>%
  split_cols_by("ARM") %>%
  analyze("COUNTRY")

build_table(lyt, adsl)
```


```{r}
# `Tplyr`
tplyr_table(adsl, ARM) %>%
  add_layer(
    group_count(COUNTRY)
  ) %>%
  build() 
```


```{r}
# tables
tabular(COUNTRY ~ ARM, data = adsl)
```

Note the difference in code required between specialized functions (`table`), formula based frameworks (`xtabs`,
`tables`), and grammar based frameworks (`dplyr`, `rtables`, `Tplyr`). Next we look at more complex tables used in
Pharma.

## Demographic Table

* N=xx
* Total Column
* Variable Labels

```{r}
lyt <- basic_table() %>%
   split_cols_by("ARM") %>%
   add_colcounts() %>%
   add_overall_col("All Patients") %>%
   analyze(c("SEX", "AGE"), afun = function(x) {
      if (is.numeric(x)) {
         in_rows(
            "n" = sum(is.na(x)),
            "Mean (sd)" = rcell(c(mean(x), sd(x)), format = "xx.xx (xx.xx)"), 
            "Median" = median(x),
            "Min - Max" = rcell(range(x), "xx.xx - xx.xx")
         )
      } else {
         in_rows(.list = as.list(table(x)))
      }
   })

build_table(lyt, adsl)
```

## Adverse Events Table

## Disposition Table

## Statistical Model Summary Table

* `broom`

## Example of Structured Data Derivation

e.g.  Comparing Biomarker Evaluable Population vs All

* nesting & splits
* Total Columns

# Analysis functions

* analysis functions 
   - same for every cell
   - different per row group
   - different per column

## Rounding Rules
This might fit into the next paper.

* Round half up
  - SAS, Excel
* R rounds to even.
  - 3.* to 4.* difference?
  - sprintf, formatC
* Data precision based rounding


# Conclusions


# Ressources

This section will be deleted, but the listed documents are useful to build the paper.

- https://github.com/RConsortium/rtrs-wg/blob/main/Minutes/RTRS_Minutes_2021-05-06.Rmd
- https://github.com/RConsortium/rtrs-wg/blob/main/Requirements/working-requirements/requirements.md
- https://roche.github.io/rtables/articles/tabulation_concepts.html
- https://roche.github.io/rtables/articles/tabulation_dplyr.html
- https://phuse.s3.eu-central-1.amazonaws.com/Deliverables/Standard+Analyses+and+Code+Sharing/Analyses+%26+Displays+Associated+with+Demographics,+Disposition+and+Medication+in+Phase+2-4+Clinical+Trials+and+Integrated+Summary+Documents.pdf
- https://advance.phuse.global/display/WEL/Safety+Analytics
- https://phuse.global/Deliverables/1
