# Overview table R packages {#tablepkgs}

We summarize various `R` packages (ordered alphabetically) to 
create production-ready tables in clinical research. 
The list is by no means complete; if you know another package that has been useful for you to
create tables or listing using clinical trial data. 
Please contribute to this chapter.


## gt

The `gt` package [@R-gt, [website](https://gt.rstudio.com/)] provides
a high-level and declarative interface for composing tables. 
Contains a variety of `formatter` functions for transforming cell values
from an input data frame to the desired reporting values (combines mutation and
aesthetic styling with options for: scientific notation, uncertainty, ranges,
percentages, suffixes, localized currency, etc.). Multiple output formats are
supported with the same declarative interface (e.g., HTML, LaTeX/PDF, and RTF).

## r2rtf

The `r2rtf` package [@R-r2rtf, [website](https://merck.github.io/r2rtf/)]
creates production-ready tables and figures in RTF format. 
The R package is designed to

- provide simple “verb” functions that correspond to each component of a table, to help you translate data frame(s) to a table in RTF file.
- enables pipes (%>%).
- only focus on table format.
- Data manipulation and analysis 
shall be handled by other R packages. (e.g., tidyverse)
minimizes package dependency

Chapter 1-9 of [The R for Clinical Study Reports and Submission](https://r4csr.org/) book 
provides examples by using `r2rtf` package. 


## rtables

The `rtables` package [@R-rtables] defines a pipe-able grammar for declaring complex table
structure layouts via nested splitting - in row- and column-space - and the
specification of analysis functions which will calculate cell values within that
structure. These pre-data table layouts are then applied to data to build the
table, during which all necessary data grouping implied by the row and column
splits, cell value calculation, and tabulation occurs automatically. Additional
table features such as titles, footers, and referential footnotes are supported.
ASCII, HTML, and PDF are supported as output formats for rendering tables.

`rtables` source code is on [GitHub](https://github.com/roche/rtables) as well
as the [documentation](https://roche.github.io/rtables/). `rtables` is also available
on [CRAN](https://CRAN.R-project.org/package=rtables).

## flextable

`flextable` [@R-flextable] is providing a *grammar* for creating and customizing tables. The
following formats are supported: 'Word' (*.docx), 'PowerPoint' (*.pptx), 'PDF'
,'HTML' and R 'Grid Graphics'. The syntax is the same for the user regardless of
the type of output to be produced. A set of functions allows the creation,
definition of cell arrangement, addition of headers or footers, formatting and
definition of cell content (i.e. text and or images). The package also offers a
set of high-level functions that allow, for example, tabular reporting of
statistical models and the creation of complex cross tabulations.

Source code is on [GitHub](https://github.com/davidgohel/flextable), a 
[user manual](https://ardata-fr.github.io/flextable-book/) is available. 
`flextable` is available on [CRAN](https://CRAN.R-project.org/package=flextable).

## tables

The `tables` package [@R-tables] provides a formula-driven interface for
computing the contents of tables and formatting them.  It was 
inspired by `SAS PROC TABULATE`, but is not compatible with it.

The user computes a table object by specifying a formula, with the
left-hand
side giving the rows, and the right-hand side giving the columns;
the formula describes the summary functions to apply and how to
organize them.  The objects can be subsetted or combined
using matrix-like operations.  Tables can
be rendered in plain text, LaTeX code to appear
in a PDF document, or HTML code for a web document.

The package is on CRAN.  Source is maintained on R-forge
at https://r-forge.r-project.org/projects/tables/ and mirrored on
Github at https://github.com/rforge/tables/.  Vignettes in the package
serve as a user manual.  Browse them at https://rdrr.io/cran/tables/, or install the package, then run
`browseVignettes(package = "tables")`.

```{r echo = FALSE}
if (!requireNamespace("tables", quietly = TRUE) ||
    packageVersion("tables") < "0.9.10")
  stop("This document requires tables version 0.9.10 or higher.")
```

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  "gt", "r2rtf", "rtables", "flextable", "tables"
), 'packages.bib')
```
