
# Formatting and Rendering Tables

Table generation usually is a two step process

1. Derive the cell value and tabulate them.
1. Create the final table output, save it to a file to be shared with collaborators. 

Chapter [Commonly Used Tables] focuses with the work involved in 1, in this chapter we
discuss the various aspects of creating the final outputs that is commonly stored
in a file with a particular file format (`pdf`, `txt`, `html`, `docx` or `rtf`).

## Title & Footnotes

Commonly rendered tables that are reported to the health authorities have titles
and footnotes with information such as:

- what is summarized in the table
- database lock date
- patient sub-population
- notes by study team
- notes regarding statistical algorithms chosen
- provenance information including path to program and when the table was created

Often footnotes include cell references.

### Title and Footnotes in rtables

The `basic_table` function in `rtables` has the argument `titles`, `subtitles`,
`main_footer`, `prov_footer` to add titles and footnotes to tables. `rtables`
also supports referential footnotes.

So for example a basic demographics table created with `rtables` via `tern` with
title and footnotes would look as follows:

```{r, comment=NA }
resetSession()
library(rtables)

lyt <- basic_table(
  title = "Demographic Table - All Patients",
  subtitles = c("Cutoff Date: June 01, 2022", "Arm B received a placebo."),
  main_footer = c("Missing data is omitted.")
) |>
  split_cols_by("ARM") |>
  analyze(c("AGE", "SEX"))

build_table(lyt, adsl)
```

### Title and Footnotes in flextable

Titles and notes can be added and formatted with the 'flextable' package. It is possible to add them in the header and in the footer. Several methods are possible but for most needs, the `add_header_lines()` and `add_footer_lines()` functions will be the easiest to use.

Let's create first a flextable from an aggregation that 
will be used to illustrate the features.

```{r message=FALSE}
resetSession()
library(flextable)
library(dplyr)

z <- adsl |> 
  group_by(ARM, SEX) |>
  summarise(avg = mean(AGE), sd = sd(AGE)) |>
  tabulator(rows = "SEX", columns = "ARM",
    Y = as_paragraph(avg, " (", sd, ")")) |> 
  as_flextable()
z
```

The following shows how to add titles or notes:

```{r}
z |>
  add_header_lines("hello world") |>
  add_footer_lines("bye bye world")
```

For Word output, users can prepend a table number that will auto-incremente.

```{r}
docx_file <- "reports/flextable-title-01.docx"
ft <- add_header_lines(z, "hello world") |>
  prepend_chunks(
    i = 1, j = 1, part = "header", 
    as_chunk("Table "), as_word_field("SEQ tab \u005C* Arabic"), 
    as_chunk(": ")) |>
  add_footer_lines("bye bye world") |>
  theme_vanilla()
save_as_docx(ft, ft, path = docx_file)
```


```{r echo=FALSE, link_preview=TRUE, path_to_doc = "reports/flextable-title-01.docx"}
```

Footnotes are also available in 'flextable' with function `footnote()`. 
The function lets users add footnotes and references to it on the table.


```{r}
footnote(z, i = c(1, 2, 2), j = c(1, 5, 7), 
         value = as_paragraph("hello world"), ref_symbols = "(1)")
```

### Titles and footnotes in the tables package

The `tables` package concentrates on the table itself.  The titles
are generally written as part of the surrounding document.  Footnotes
would be added after constructing the table by modifying 
individual entries.

Alternatively, only the footnote markers need to be added by modifying entries, and then the
table object is converted to the format 
supported by the `kableExtra` package, and customizations 
can then be applied.

```{r}
resetSession()
adsl <- cadsl

library(tables)
library(kableExtra)

sd_in_parens <- function(x) sprintf("(%.1f)", sd(x))

tab <- tabular(SEX ~ Heading()*ARM*
                     Heading()*AGE*
                     Heading()*(mean + sd_in_parens), 
               data = adsl)

rowLabels(tab)[1,1] <- paste(rowLabels(tab)[1,1], "<sup>a</sup>")
tab[2,2] <- sprintf("%s%s", tab[2,2], "<sup>b</sup>")
tab[2,3] <- sprintf("%.2f%s", tab[2,3], "<sup>b</sup>")

toKable(tab) |>
  kable_styling(full_width = FALSE) |>
  add_footnote(c("This is a label footnote", "These are cell footnotes"))
```

## Captions

A caption is a single paragraph of text describing the table it
is associated to. Captions are often used because they allow you to
cross-reference tables or list them in a 'list of tables' with the corresponding
page numbers.

### Captions in flextable

The `set_caption()` function in 'flextable' is the recommanded way to add
captions. 

```{r}
resetSession()
library(flextable)

flextable(head(cars)) |>
  set_caption(
    caption = "a caption",
    autonum = officer::run_autonum(seq_id = "tab", bkm = "flextable-label"))
```

In bookdown, use the syntax `\@ref(tab:flextable-label)` to create a linked 
reference to the table. See an example of a reference: \@ref(tab:flextable-label).

With 'Quarto', the R chunk code should be transformed as: 

````r
#| label: tbl-flextable-label
#| tbl-cap: a caption
flextable(head(cars))
````

### Captions in the tables package

As with titles, captions would be added as part of the
surrounding document rather than part of the table object.

## Pagination

Historically tables have been printed to paper for submissions. Hence large tables
that would not fit onto a single printed page (e.g. letter & portrait) would have to 
be split into multiple tables that can be printed to the preferred page size. This 
process of splitting the table is called *pagination* of tables.

Pagination is a complex task as it requires to repeat information so that the tables
on each page are correct tables with all relevant information for context.

### Pagination in rtables

In `rtables` pagination can be done with the `paginate_table` function which has 
a number of arguments to split the table, see the [example documentation](https://roche.github.io/rtables/reference/paginate.html).

### Pagination in flextable

In 'flextable', there is no concept of pagination. The recommended method is to
create a flextable for each modality of a group and to separate them with a page
break; from R, this implies looping over a set of modalities with a `for`
statement or a call to the `lapply()` function. 

Note this concept of page break is only valid for paginated documents, i.e.
Word, PDF, eventually HTML with
[paged.js](https://pagedjs.org/made-with-paged.js.html) formats. 

In an 'R Markdown' document, the markdown code to insert a page break is
`\newpage`; with the package 'officer', you have to use `body_add_break()` (for
Word documents only).

### Pagination in the tables package

Tables that span multiple pages are supported in PDF output using
the LaTeX `longtable` package, specified when converting to LaTeX.



## Rendering Tables

The choice of file format is often dictated by your company's processes to 
include the tables in a report.

### ASCII

### PDF

### HTML

### WORD

### RTF


