---
output: html_document
editor_options: 
  chunk_output_type: console
---

## Demographic Tables

### rtables

Using `rtables` only:

```{r, comment= NA}
resetSession()
library(rtables)

a_demo_num <- function(x) {
    in_rows(n = length(x),
            "Mean (SD)" = rcell(c(mean(x, na.rm = TRUE),
                                  sd(x, na.rm=TRUE)), format = "xx.x (xx.x)"),
            "Median" = median(x,na.rm = TRUE),
            "Min - Max" = rcell(range(x, na.rm = TRUE), format = "xx.x - xx.x"))
}

a_demo_fac <- function(x) {
    in_rows(.list = c(c(n = length(x)), table(x)))
}

lyt <- basic_table(title = "x.x: Study Subject Data",
                   subtitles= c("x.x.x: Demographic Characteristics",
                                "Table x.x.x.x: Demographic Characteristics - Full Analysis Set"),
                   prov_footer = "Source: ADSL DDMMYYYY hh:mm; Listing x.xx; SDTM package: DDMMYYYY") |>
  split_cols_by("ARM") |>
  analyze(c("AGE", "SEX", "COUNTRY"), afun = list(AGE = a_demo_num, SEX = a_demo_fac,
                                                  COUNTRY = a_demo_fac))
 
build_table(lyt, ex_adsl)
```

Using `rtables` and `tern`

```{r, comment=NA}
library(tern)
lyt <- basic_table(title = "x.x: Study Subject Data",
                   subtitles= c("x.x.x: Demographic Characteristics",
                                "Table x.x.x.x: Demographic Characteristics - Full Analysis Set"),
                   prov_footer = "Source: ADSL DDMMYYYY hh:mm; Listing x.xx; SDTM package: DDMMYYYY") |>
  split_cols_by("ARM") |>
  summarize_vars(c("AGE", "SEX", "COUNTRY"))

build_table(lyt, ex_adsl)
```


### gt

```{r, comment= NA}
resetSession()
library(gt)
library(tidyverse)

# We will use ex_adsl but will assign a unit to the Age column

ex_adsl <- formatters::ex_adsl
gt_adsl <- ex_adsl
attr(gt_adsl$AGE, "units") <- "Years"

# This is a customized summary function
# It creates numeric and categorical summaries for specified variables, following the rtables exmaple

custom_summary <- function(df, group_var, sum_var){
  group_var <- rlang::ensym(group_var)
  sum_var <- rlang::ensym(sum_var)
  
  is_categorical <- is.character(eval(expr(`$`(df, !!sum_var)))) | is.factor(eval(expr(`$`(df, !!sum_var)))) 
  
  if (is_categorical){
    df <- df  |>  
      dplyr::group_by(!!group_var) |> 
      dplyr::mutate(N = n()) |> 
      dplyr::ungroup() |> 
      dplyr::group_by(!!group_var, !!sum_var) |> 
      dplyr::summarize(
        val = n(),
        sd = 100*n()/mean(N),
        .groups = "drop"
        ) |>
      tidyr::pivot_wider(id_cols = !!sum_var, names_from = !!group_var, values_from = c(val, sd)) |> 
      dplyr::rename(label = !!sum_var) |> 
      dplyr::mutate(isnum = FALSE,
                    across(where(is.numeric), ~ifelse(is.na(.), 0, .))) 
    
    sum_unit <- ", n (%)"
    
  } else {
    
    sum_unit <- sprintf(" (%s)", attr(eval(expr(`$`(df, !!sum_var))), "units"))
    
    df <- df |> 
      dplyr::group_by(!!group_var) |> 
      dplyr::summarize(
        n = sum(!is.na(!!sum_var)),
        mean = mean(!!sum_var, na.rm = TRUE),
        sd = sd(!!sum_var, na.rm = TRUE),
        median = median(!!sum_var, na.rm = TRUE),
        min = min(!!sum_var, na.rm = TRUE),
        max = max(!!sum_var, na.rm = TRUE),
        min_max = NA,
        .groups = "drop"
        ) |> 
      tidyr::pivot_longer(cols = c(n, mean, median, min_max), names_to = "label", values_to = "val") |> 
      dplyr::mutate(sd = ifelse(label == "mean", sd, NA),
                max = ifelse(label == "min_max", max, NA),
                min = ifelse(label == "min_max", min, NA),
                label = dplyr::recode(label, "mean" = "Mean (SD)", "min_max" = "Min - Max", "median" = "Median")) |>  
      tidyr::pivot_wider(id_cols = label, names_from = !!group_var, values_from = c(val, sd, min, max)) |> 
      dplyr::mutate(isnum = TRUE)
    
  }
  
  df |> 
    dplyr::mutate(category = paste0(stringr::str_to_title(deparse(substitute(!!sum_var))),
                                     sum_unit)) 
}

# Perform aggregation for variables Age, Sex and Country

adsl_summary <- purrr::map_df(.x = vars(AGE, SEX, COUNTRY),
                             .f = ~custom_summary(df = gt_adsl, group_var = ARM, sum_var = !!.x)) 

# Count number of patients per Arm

total_n_arm <- ex_adsl |>
  dplyr::group_by(ARM) |> 
  dplyr::summarise(N = n(), .groups = "drop")

# gt

gt(adsl_summary, 
   rowname_col = "label",
   groupname_col = "category") |>
  tab_header(
    title = "x.x: Study Subject Data",
    subtitle = md("x.x.x: Demographic Characteristics  \n  Table x.x.x.x: Demographic Characteristics - Full Analysis Set"),
    preheader = c("Protocol: XXXXX", "Cutoff date: DDMMYYYY")
  ) |> 
  tab_source_note("Source: ADSL DDMMYYYY hh:mm; Listing x.xx; SDTM package: DDMMYYYY") |> 
  opt_align_table_header(align = "left") |> 
  fmt_integer(columns = starts_with(c("val", "min", "max")), rows = label != "Mean (SD)") |> 
  fmt_number(columns = starts_with(c("val", "sd")), rows = label == "Mean (SD)", decimals = 1) |> 
  fmt_number(columns = starts_with("sd"), rows = isnum == FALSE, decimals = 1) |> 
  sub_missing(missing_text = "") |> 
  summary_rows(
    groups = c("Sex, n (%)", "Country, n (%)"),
    columns = starts_with("val"),
    fns = list(n = ~sum(.)),
    missing_text = "",
    formatter = fmt_integer #,
    # side = "top"
    ) |> 
  cols_merge_n_pct(col_n = "val_A: Drug X", col_pct = "sd_A: Drug X") |>
  cols_merge_n_pct(col_n = "val_B: Placebo", col_pct = "sd_B: Placebo") |> 
  cols_merge_n_pct(col_n = "val_C: Combination", col_pct = "sd_C: Combination") |> 
  cols_merge_range(col_begin = "min_A: Drug X", col_end = "max_A: Drug X", sep = " - ") |> 
  cols_merge_range(col_begin = "min_B: Placebo", col_end = "max_B: Placebo", sep = " - ") |> 
  cols_merge_range(col_begin = "min_C: Combination", col_end = "max_C: Combination", sep = " - ") |> 
  cols_merge(columns = c("val_A: Drug X", "min_A: Drug X"), pattern = "{1}{2}") |> 
  cols_merge(columns = c("val_B: Placebo", "min_B: Placebo"), pattern = "{1}{2}") |> 
  cols_merge(columns = c("val_C: Combination", "min_C: Combination"), pattern = "{1}{2}") |> 
  cols_hide(columns = isnum) |> 
  cols_align(align = c("center"),
             columns = c("val_A: Drug X", "val_B: Placebo", "val_C: Combination")) |> 
  cols_align(align = "center",
             columns = 1) |> 
  cols_label("val_A: Drug X" = md(sprintf("A: Drug X  \n  N=%i (100%%)", total_n_arm$N[1])),
             "val_B: Placebo" = md(sprintf("B: Placebo  \n  N=%i (100%%)", total_n_arm$N[2])),
             "val_C: Combination" = md(sprintf("C: Combination  \n  N=%i (100%%)", total_n_arm$N[3]))) |> 
  tab_options(
    table.font.size = 9,
    page.orientation = "landscape",
    page.numbering = TRUE,
    page.header.use_tbl_headings = TRUE,
    page.footer.use_tbl_notes = TRUE)
```

### flextable



```{r}
# The two steps in creating 'Demographic Tables' are:
# 
# - summarize the information with the `flextable::summarizor()` function. 
# It computes a set of statistics for each variable by groups. It returns 
# a data.frame ready to be used by `flextable::as_flextable()`.
# - Create the flextable with the `as_flextable()` function.

resetSession()
ex_adsl <- formatters::ex_adsl

library(flextable)
library(tidyverse)
library(officer)

set_flextable_defaults(
  border.color = "#AAAAAA", font.family = "Open Sans",
font.size = 10, padding = 3, line_spacing = 1.4
)

# data
adsl <- select(ex_adsl, AGE, SEX, COUNTRY, ARM)

# In the illustration, we use labels from the column attributes.  

col_labels <- map_chr(adsl, function(x) attr(x, "label"))

# Now let's use the labels and customize the ‘flextable’ output.

ft <- summarizor(adsl, by = "ARM") |>
  as_flextable(sep_w = 0, separate_with = "variable", 
               spread_first_col = TRUE) |>
  align(i = ~ !is.na(variable), align = "left") |> 
  prepend_chunks(i = ~ is.na(variable), j  ="stat", as_chunk("\t") ) |> 
  labelizor(j = c("stat"), 
            labels = col_labels, part = "all") |> 
  autofit() |>
  add_header_lines(
    c("x.x: Study Subject Data",
      "x.x.x: Demographic Characteristics",
      "Table x.x.x.x: Demographic Characteristics - Full Analysis Set")) |> 
  add_footer_lines("Source: ADSL DDMMYYYY hh:mm; Listing x.xx; SDTM package: DDMMYYYY")

ft
```

### tables

The `tables` package uses a different style
than the other packages for tables such as this, where
there are separate sections for age, sex and country
breakdowns of the data.  Rather than putting
the section heading on a separate line, it normally
puts the heading in a separate column to the left
of the other columns.

```{r}
resetSession()

ex_adsl <- formatters::ex_adsl

library(tables)
table_options(doCSS = TRUE)

meansd <- function(x) sprintf("%.1f (%.1f)", mean(x), sd(x))

iqr <- function(x) quantile(x, 0.75) - quantile(x, 0.25)

medianiqr <- function(x) sprintf("%.1f (%.1f)", median(x), iqr(x))

minmax <- function(x) sprintf("%.1f - %.1f", min(x), max(x))

countpercent <- function(num, denom) 
  sprintf("%d (%.1f%%)", 
          length(num), 
          100*length(num)/length(denom))

count <- function(x) sprintf("(N=%d)", length(x))

tab <- tabular( Heading()*1*Heading()*count +
         Heading("Age (Years)")*
           AGE * (Heading("Mean (SD)")*meansd +
                  Heading("Median (IQR)")*medianiqr +
                  Heading("Min - Max")*minmax) +
         (Heading("Sex, n, (%)")*SEX +
          Heading("Country, n, (%)")*COUNTRY)*
            Heading()*Percent(denom = Equal(ARM), fn = countpercent) ~ 
         Heading()*ARM, 
         data = ex_adsl )

useGroupLabels(tab, indent = "&emsp;")
```

### tidytlg

```{r, warning=FALSE}
resetSession()
library(dplyr)
library(tidytlg)

adsl <- formatters::ex_adsl

# create univariate stats for age
tbl1 <- univar(adsl,
               rowvar = "AGE",
               colvar = "ARM",
               statlist = statlist(c("N","MEANSD","MEDIAN","RANGE")),
               row_header = "Age (years)",
               decimal = 0)

# create counts (percentages) for gender categories
tbl2 <- freq(adsl,
             rowvar = "SEX",
             colvar = "ARM",
             statlist = statlist(c("N", "n (x.x%)")),
             row_header = "Gender, n(%)")

# create counts (percentages) for country
tbl3 <- freq(adsl,
             rowvar = "COUNTRY",
             colvar = "ARM",
             statlist = statlist(c("N", "n (x.x%)")),
             row_header = "Country, n(%)",
             descending_by = "C: Combination")

# combine analysis results together
tbl <- bind_table(tbl1, tbl2, tbl3)

# output the analysis results
gentlg(huxme       = tbl,
       format      = "HTML",
       print.hux = FALSE,
       file        = "Table x.x.x.x",
       orientation = "portrait",
       title = "Demographic Characteristics - Full Analysis Set",
       footers = "Source: ADSL DDMMYYYY hh:mm; Listing x.xx; SDTM package: DDMMYYYY",
       colheader = c("","A: Drug X","B: Placebo","C: Combination"))

```

### tfrmt

Please note that the`tfrmt` package is intended for use with mock data or ARD (analysis results data).
This package creates the same tables as other packages but requires the starting data to be transformed first. 

The first chunk of code takes the CDISC data and modifies it into an ARD. The second chunk demonstrates 
how to format the table.

```{r}
resetSession()
library(tidyverse)
library(tfrmt)

# Get data
data("cadsl", package = "random.cdisc.data")

# Number of unique subjects per ARM
big_n <- cadsl |>
  dplyr::group_by(ARM) |>
  dplyr::summarize(
    N = dplyr::n_distinct(USUBJID)
  )

# Join big_n with adsl
adsl_with_n <- cadsl |>
  dplyr::left_join(big_n, by = "ARM")

# Explore column: AGE
age_stats <-
  adsl_with_n |>
  group_by(ARM) |>
  reframe(
    n = n_distinct(USUBJID),
    Mean = mean(AGE),
    SD = sd(AGE),
    Median = median(AGE),
    Min = min(AGE),
    Max = max(AGE)
  ) |>
  pivot_longer(
    c("n", "Mean", "SD", "Median", "Min", "Max")
  ) |>
  mutate(
    group = "Age (years)",
    label = case_when(name == "Mean" ~ "Mean (SD)",
                      name == "SD" ~ "Mean (SD)",
                      name == "Min" ~ "Min - Max",
                      name == "Max" ~ "Min - Max",
                      TRUE ~ name)
  )

sex_n <-
  adsl_with_n |>
  group_by(ARM, SEX) |>
  reframe(
    n = n(),
    pct = (n/N)*100
  ) |>
  distinct() |>
  pivot_longer(
    c("n", "pct")
  ) |>
  rename(
    label = SEX
  ) |>
  mutate(
    group = "Sex"
  )

# Explore column: COUNTRY
country_n <-
  adsl_with_n |>
  group_by(ARM, COUNTRY) |>
  reframe(
    n = n(),
    pct = (n/N)*100
) |>
  distinct() |>
  pivot_longer(
    c("n", "pct")
  ) |>
  rename(
    label = COUNTRY
  ) |>
  mutate(
    group = "Country"
  )

# Header n
header_n <- big_n |>
  dplyr::rename(value = N) |>
  dplyr::mutate(name = "header_n")

# Create ARD
demog_ard <-
  bind_rows(
    age_stats,
    sex_n,
    country_n,
    #header_n
  ) |>
  rename(
    column = ARM,
    param = name
  ) |>
  select(
    group, label, param, column, value
  ) |>
  group_by(group, label)
```

Now we can used the `demog_ard` to make the demographic table using `tfrmt`.

```{r}
tfrmt(
  # Add titles
  title = "x.x: Study Subject Data",
  subtitle = c("x.x.x: Demographic Characteristics. \n
               Table x.x.x.x: Demographic Characteristics - Full Analysis Set"),

  # Specify table features
  group = group,
  label = label,
  column = column,
  param = param,
  value = value,

  # Define cell formatting
  body_plan = body_plan(
    # Define rounding and structure of values in each row
    frmt_structure(group_val = ".default", label_val = ".default", frmt("xx")),

    frmt_structure(group_val = "Age (years)",
                   label_val = c("Mean (SD)"),
                   frmt_combine(
                     "{Mean} ({SD})",
                     Mean = frmt("xx.x"),
                     SD = frmt("x.x") )),

    frmt_structure(group_val = "Age (years)",
                   label_val = c("Min - Max"),
                   frmt_combine(
                     "{Min} - {Max}",
                     frmt("xx.x") )),

    frmt_structure(group_val = "Sex",
                   label_val = c("M", "F", "U", "UNDIFFERENTIATED"),
                   frmt_combine(
                     "{n} ({pct}%)",
                     n = frmt("XXX"),
                     pct = frmt("XX.X") )),

    frmt_structure(group_val = "Country",
                   label_val = c("CHN", "USA", "BRA", "PAK", "NGA", "RUS", "JPN", "GBR", "CAN", "NA"),
                   frmt_combine(
                     "{n} ({pct}%)",
                     n = frmt("XXX"),
                     pct = frmt("XX.X") ))
  ),

  # Align values on decimal places and spaces
  col_style_plan = col_style_plan(
    col_style_structure(col = matches("[A-Z]:.*"),
                        align = c(".", " "))
  ) ) %>%
  print_to_gt(demog_ard)
```

See this vignette for more details on formatting functions: [link to website](https://gsk-biostatistics.github.io/tfrmt/articles/body_plan.html)

See this vignette for the completed table example: [link to website](https://gsk-biostatistics.github.io/tfrmt/articles/examples.html)

