---
output: html_document
editor_options: 
  chunk_output_type: console
---

## Demographic Tables

### rtables

Using `rtables` only:

```{r, comment= NA}
resetSession()
library(rtables)

a_demo_num <- function(x) {
    in_rows(n = length(x),
            "Mean (SD)" = rcell(c(mean(x, na.rm = TRUE),
                                  sd(x, na.rm=TRUE)), format = "xx.x (xx.x)"),
            "Median" = median(x,na.rm = TRUE),
            "Min - Max" = rcell(range(x, na.rm = TRUE), format = "xx.x - xx.x"))
}

a_demo_fac <- function(x) {
    in_rows(.list = c(c(n = length(x)), table(x)))
}

lyt <- basic_table(title = "x.x: Study Subject Data",
                   subtitles= c("x.x.x: Demographic Characteristics",
                                "Table x.x.x.x: Demographic Characteristics - Full Analysis Set"),
                   prov_footer = "Source: ADSL DDMMYYYY hh:mm; Listing x.xx; SDTM package: DDMMYYYY") |>
  split_cols_by("ARM") |>
  analyze(c("AGE", "SEX", "COUNTRY"), afun = list(AGE = a_demo_num, SEX = a_demo_fac,
                                                  COUNTRY = a_demo_fac))
 
build_table(lyt, ex_adsl)
```

Using `rtables` and `tern`

```{r, comment=NA}
library(tern)
lyt <- basic_table(title = "x.x: Study Subject Data",
                   subtitles= c("x.x.x: Demographic Characteristics",
                                "Table x.x.x.x: Demographic Characteristics - Full Analysis Set"),
                   prov_footer = "Source: ADSL DDMMYYYY hh:mm; Listing x.xx; SDTM package: DDMMYYYY") |>
  split_cols_by("ARM") |>
  summarize_vars(c("AGE", "SEX", "COUNTRY"))

build_table(lyt, ex_adsl)
```


### gt

```{r, comment= NA}
resetSession()
library(gt)
library(tidyverse)

# We will use ex_adsl but will assign a unit to the Age column

ex_adsl <- formatters::ex_adsl
gt_adsl <- ex_adsl
attr(gt_adsl$AGE, "units") <- "Years"

# This is a customized summary function
# It creates numeric and categorical summaries for specified variables, following the rtables exmaple

custom_summary <- function(df, group_var, sum_var){
  group_var <- rlang::ensym(group_var)
  sum_var <- rlang::ensym(sum_var)
  
  is_categorical <- is.character(eval(expr(`$`(df, !!sum_var)))) | is.factor(eval(expr(`$`(df, !!sum_var)))) 
  
  if (is_categorical){
    df <- df  |>  
      dplyr::group_by(!!group_var) |> 
      dplyr::mutate(N = n()) |> 
      dplyr::ungroup() |> 
      dplyr::group_by(!!group_var, !!sum_var) |> 
      dplyr::summarize(
        val = n(),
        sd = 100*n()/mean(N),
        .groups = "drop"
        ) |>
      tidyr::pivot_wider(id_cols = !!sum_var, names_from = !!group_var, values_from = c(val, sd)) |> 
      dplyr::rename(label = !!sum_var) |> 
      dplyr::mutate(isnum = FALSE,
                    across(where(is.numeric), ~ifelse(is.na(.), 0, .))) 
    
    sum_unit <- ", n (%)"
    
  } else {
    
    sum_unit <- sprintf(" (%s)", attr(eval(expr(`$`(df, !!sum_var))), "units"))
    
    df <- df |> 
      dplyr::group_by(!!group_var) |> 
      dplyr::summarize(
        n = sum(!is.na(!!sum_var)),
        mean = mean(!!sum_var, na.rm = TRUE),
        sd = sd(!!sum_var, na.rm = TRUE),
        median = median(!!sum_var, na.rm = TRUE),
        min = min(!!sum_var, na.rm = TRUE),
        max = max(!!sum_var, na.rm = TRUE),
        min_max = NA,
        .groups = "drop"
        ) |> 
      tidyr::pivot_longer(cols = c(n, mean, median, min_max), names_to = "label", values_to = "val") |> 
      dplyr::mutate(sd = ifelse(label == "mean", sd, NA),
                max = ifelse(label == "min_max", max, NA),
                min = ifelse(label == "min_max", min, NA),
                label = dplyr::recode(label, "mean" = "Mean (SD)", "min_max" = "Min - Max", "median" = "Median")) |>  
      tidyr::pivot_wider(id_cols = label, names_from = !!group_var, values_from = c(val, sd, min, max)) |> 
      dplyr::mutate(isnum = TRUE)
    
  }
  
  df |> 
    dplyr::mutate(category = paste0(stringr::str_to_title(deparse(substitute(!!sum_var))),
                                     sum_unit)) 
}

# Perform aggregation for variables Age, Sex and Country

adsl_summary <- purrr::map_df(.x = vars(AGE, SEX, COUNTRY),
                             .f = ~custom_summary(df = gt_adsl, group_var = ARM, sum_var = !!.x)) 

# Count number of patients per Arm

total_n_arm <- ex_adsl |>
  dplyr::group_by(ARM) |> 
  dplyr::summarise(N = n(), .groups = "drop")

# gt

gt(adsl_summary, 
   rowname_col = "label",
   groupname_col = "category") |>
  tab_header(
    title = "x.x: Study Subject Data",
    subtitle = md("x.x.x: Demographic Characteristics  \n  Table x.x.x.x: Demographic Characteristics - Full Analysis Set"),
    preheader = c("Protocol: XXXXX", "Cutoff date: DDMMYYYY")
  ) |> 
  tab_source_note("Source: ADSL DDMMYYYY hh:mm; Listing x.xx; SDTM package: DDMMYYYY") |> 
  opt_align_table_header(align = "left") |> 
  fmt_integer(columns = starts_with(c("val", "min", "max")), rows = label != "Mean (SD)") |> 
  fmt_number(columns = starts_with(c("val", "sd")), rows = label == "Mean (SD)", decimals = 1) |> 
  fmt_number(columns = starts_with("sd"), rows = isnum == FALSE, decimals = 1) |> 
  sub_missing(missing_text = "") |> 
  summary_rows(
    groups = c("Sex, n (%)", "Country, n (%)"),
    columns = starts_with("val"),
    fns = list(n = ~sum(.)),
    missing_text = "",
    formatter = fmt_integer #,
    # side = "top"
    ) |> 
  cols_merge_n_pct(col_n = "val_A: Drug X", col_pct = "sd_A: Drug X") |>
  cols_merge_n_pct(col_n = "val_B: Placebo", col_pct = "sd_B: Placebo") |> 
  cols_merge_n_pct(col_n = "val_C: Combination", col_pct = "sd_C: Combination") |> 
  cols_merge_range(col_begin = "min_A: Drug X", col_end = "max_A: Drug X", sep = " - ") |> 
  cols_merge_range(col_begin = "min_B: Placebo", col_end = "max_B: Placebo", sep = " - ") |> 
  cols_merge_range(col_begin = "min_C: Combination", col_end = "max_C: Combination", sep = " - ") |> 
  cols_merge(columns = c("val_A: Drug X", "min_A: Drug X"), pattern = "{1}{2}") |> 
  cols_merge(columns = c("val_B: Placebo", "min_B: Placebo"), pattern = "{1}{2}") |> 
  cols_merge(columns = c("val_C: Combination", "min_C: Combination"), pattern = "{1}{2}") |> 
  cols_hide(columns = isnum) |> 
  cols_align(align = c("center"),
             columns = c("val_A: Drug X", "val_B: Placebo", "val_C: Combination")) |> 
  cols_align(align = "center",
             columns = 1) |> 
  cols_label("val_A: Drug X" = md(sprintf("A: Drug X  \n  N=%i (100%%)", total_n_arm$N[1])),
             "val_B: Placebo" = md(sprintf("B: Placebo  \n  N=%i (100%%)", total_n_arm$N[2])),
             "val_C: Combination" = md(sprintf("C: Combination  \n  N=%i (100%%)", total_n_arm$N[3]))) |> 
  tab_options(
    table.font.size = 9,
    page.orientation = "landscape",
    page.numbering = TRUE,
    page.header.use_tbl_headings = TRUE,
    page.footer.use_tbl_notes = TRUE)
```

### flextable



```{r}
# The two steps in creating 'Demographic Tables' are:
# 
# - summarize the information with the `flextable::summarizor()` function. 
# It computes a set of statistics for each variable by groups. It returns 
# a data.frame ready to be used by `flextable::as_flextable()`.
# - Create the flextable with the `as_flextable()` function.

resetSession()
ex_adsl <- formatters::ex_adsl

library(flextable)
library(tidyverse)
library(officer)

set_flextable_defaults(
  border.color = "#AAAAAA", font.family = "Open Sans",
font.size = 10, padding = 3, line_spacing = 1.4
)

# data
adsl <- select(ex_adsl, AGE, SEX, COUNTRY, ARM)

# In the illustration, we use labels from the column attributes.  

col_labels <- map_chr(adsl, function(x) attr(x, "label"))

# Now let's use the labels and customize the ‘flextable’ output.

ft <- summarizor(adsl, by = "ARM") |>
  as_flextable(sep_w = 0, separate_with = "variable", 
               spread_first_col = TRUE) |>
  align(i = ~ !is.na(variable), align = "left") |> 
  prepend_chunks(i = ~ is.na(variable), j  ="stat", as_chunk("\t") ) |> 
  labelizor(j = c("stat"), 
            labels = col_labels, part = "all") |> 
  autofit() |>
  add_header_lines(
    c("x.x: Study Subject Data",
      "x.x.x: Demographic Characteristics",
      "Table x.x.x.x: Demographic Characteristics - Full Analysis Set")) |> 
  add_footer_lines("Source: ADSL DDMMYYYY hh:mm; Listing x.xx; SDTM package: DDMMYYYY")

ft
```

### tables

The `tables` package uses a different style
than the other packages for tables such as this, where
there are separate sections for age, sex and country
breakdowns of the data.  Rather than putting
the section heading on a separate line, it normally
puts the heading in a separate column to the left
of the other columns.

```{r}
resetSession()

ex_adsl <- formatters::ex_adsl

library(tables)
table_options(doCSS = TRUE)

meansd <- function(x) sprintf("%.1f (%.1f)", mean(x), sd(x))

iqr <- function(x) quantile(x, 0.75) - quantile(x, 0.25)

medianiqr <- function(x) sprintf("%.1f (%.1f)", median(x), iqr(x))

minmax <- function(x) sprintf("%.1f - %.1f", min(x), max(x))

countpercent <- function(num, denom) 
  sprintf("%d (%.1f%%)", 
          length(num), 
          100*length(num)/length(denom))

count <- function(x) sprintf("(N=%d)", length(x))

tab <- tabular( Heading()*1*Heading()*count +
         Heading("Age (Years)")*
           AGE * (Heading("Mean (SD)")*meansd +
                  Heading("Median (IQR)")*medianiqr +
                  Heading("Min - Max")*minmax) +
         (Heading("Sex, n, (%)")*SEX +
          Heading("Country, n, (%)")*COUNTRY)*
            Heading()*Percent(denom = Equal(ARM), fn = countpercent) ~ 
         Heading()*ARM, 
         data = ex_adsl )

useGroupLabels(tab, indent = "&emsp;")
```
