---
output: html_document
editor_options: 
  chunk_output_type: console
---
## Concomitant Medications

### rtables

```{r}
resetSession()

library(rtables)
data("cadcm", package = "random.cdisc.data")
data("cadsl", package = "random.cdisc.data")

one_count_pct_gen <- function(label = NULL) {
    function(x, .N_col) {
       ret <- rcell(length(unique(x)) * c(1, 1/.N_col),
                    format = "xx (xx.x%)")
       if(!is.null(label))
           obj_label(ret) <- label
       ret
    }
}

lyt <- basic_table(title = "Conmed Example",
                   subtitles = "Uses the adcm dataset from random.cdisc.data",
                   show_colcounts = TRUE) %>%
    split_cols_by("ARM") %>%
    analyze("USUBJID", afun = one_count_pct_gen("At Least One Concomittant Med")) %>%
    split_rows_by("CMCLAS", split_fun = trim_levels_in_group("CMTRT")) %>%
    analyze("CMTRT", afun = function(df, .N_col) {
        cmtrtvec <- df$CMTRT
        spl_usubj <- split(df$USUBJID, cmtrtvec)
        fn <- one_count_pct_gen()
        cells <- lapply(spl_usubj, fn, .N_col = .N_col)
        names(cells) <- names(spl_usubj)
        in_rows(.list = cells)
    })

build_table(lyt, cadcm, alt_counts_df = cadsl)
```

### flextable

This is again a situation where the code required to create a flextable directly
requires too much data preparation. In the following example, we convert the
'rtables' object to a flextable using the `as_flextable()` method and then we
change its aspect. 

```{r}
library(flextable)

tt_to_flextable(build_table(lyt, cadcm, alt_counts_df = cadsl)) |> 
  theme_booktabs() |> 
  font(fontname = "Open Sans") |> 
  bold(i = ~ V2 %in% "", j = 1, bold = TRUE) |> 
  bold(i = 1, j = 1, bold = TRUE) |> 
  align(j = 2:4, align = "center", part = "all") |> 
  set_table_properties(layout = "fixed") |> 
  autofit() |> 
  mk_par(i = 1, j = 1, part = "header",
         as_paragraph(as_chunk("Conmed Example", props = fp_text_default(font.size = 14)))) |> 
  mk_par(i = 2, j = 1, part = "header",
         as_paragraph("Uses the adcm dataset from ", as_b("random.cdisc.data")))
```

### gt

```{r, comment=NA}
resetSession()

library(dplyr)
library(tidyr)
library(gt)

data("cadcm", package = "random.cdisc.data")
data("cadsl", package = "random.cdisc.data")


cmdecod_levels <- c("Number of sujects with any concomitant medication", levels(cadcm$CMDECOD))
cmclas_levels <- c(NA, levels(cadcm$CMCLAS))

adcm <- cadcm |> 
  dplyr::select(CMDECOD, CMCLAS, TRT01A) |> 
  dplyr::mutate(
    CMDECOD = factor(CMDECOD, levels = cmdecod_levels),
    CMCLAS = factor(CMCLAS, levels = cmclas_levels)
    )

ct_cm <- cadcm |> 
  dplyr::summarize(
    n = dplyr::n_distinct(USUBJID), 
    .by = TRT01A
    ) |> 
  dplyr::left_join(count(cadsl, TRT01A, name = "nall"), by = "TRT01A") |> 
  dplyr::mutate(
    pct = n / nall, nall = NULL,
    CMDECOD = factor("Number of sujects with any concomitant medication", levels = cmdecod_levels)
    )

ct_adcm <- cadcm |> 
  dplyr::summarize(
    n = dplyr::n_distinct(USUBJID), 
    .by = c(TRT01A, CMCLAS, CMDECOD)
    ) |> 
  dplyr::left_join(count(cadsl, TRT01A, name = "nall"), by = "TRT01A") |> 
  dplyr::mutate(pct = n / nall, nall = NULL)

gt_adcm <- dplyr::bind_rows(ct_cm, ct_adcm) |>  
  tidyr::pivot_wider(id_cols = c(CMCLAS, CMDECOD), names_from = TRT01A, values_from = c(n, pct)) 


trt_n <- cadsl |> 
  dplyr::filter(SAFFL == "Y") |> 
  dplyr::summarize(
    n = sprintf("%s  \n(N=%i)", unique(TRT01A), dplyr::n()), 
    .by = TRT01A
    ) 

header_n <- as.list(trt_n$n) 
names(header_n) <- paste("n", dplyr::pull(trt_n, TRT01A), sep = "_")


gt_adcm |> 
  gt(rowname_col = "CMDECOD") |> 
  tab_header(
    title = "Conmed Example",
    subtitle = md("Uses the *adcm* dataset from **random.cdisc.data**")
  ) |> 
  opt_align_table_header(align = "left") |> 
  fmt_percent(columns = dplyr::starts_with("pct_"), decimals = 1) |> 
  cols_merge_n_pct(col_n = "n_A: Drug X", col_pct = "pct_A: Drug X") |> 
  cols_merge_n_pct(col_n = "n_B: Placebo", col_pct = "pct_B: Placebo") |> 
  cols_merge_n_pct(col_n = "n_C: Combination", col_pct = "pct_C: Combination") |> 
  tab_row_group(
    label = "medcl A",
    rows = CMCLAS == "medcl A"
  ) |>
  tab_row_group(
    label = "medcl B",
    rows = CMCLAS == "medcl B"
  ) |>
  tab_row_group(
    label = "medcl C",
    rows = CMCLAS == "medcl C"
  ) |>
  row_group_order(
    groups = c(NA, paste("medcl", LETTERS[1:2])) 
  ) |> 
  cols_hide(CMCLAS) |> 
  cols_label(
    .list = header_n,
    .fn = md
  ) |> 
  cols_width(
    1 ~ px(500),
    everything() ~ px(150)
  ) |> 
  cols_align(
    align = "center",
    columns = everything()
  ) |> 
  cols_align(
    align = "left",
    columns = 1
  ) 

```

### tables

```{r}
resetSession()

data("cadcm", package = "random.cdisc.data")

library(tables)
table_options(doCSS = TRUE)

subject_counts <- table(adsl$ARM)

countpercentid <- function(num, ARM) {
  n <- length(unique(num))
  if (n == 0) pct <- 0
  else        pct <- 100*n/subject_counts[ARM[1]]
  sprintf("%d (%.2f%%)", 
          length(unique(num)), 
          pct)
}

count <- function(x) sprintf("(N=%d)", length(x))

heading <- tabular(Heading("")*1*Heading("")*count ~ 
                   Heading()*ARM, 
                   data = adsl)

body <- tabular( (Heading("Any concomitant medication")*1*Heading("")*1 + 
                  Heading()*CMCLAS*
                    Heading()*CMDECOD*DropEmpty(which = "row"))*
                 Heading()*countpercentid*Arguments(ARM = TRT01A)*
                   Heading()*USUBJID ~
                 Heading()*TRT01A, 
                 data = cadcm)

useGroupLabels(rbind(heading, body), indent = "&emsp;")
```

### tidytlg

```{r}
resetSession()
library(dplyr)
library(tidytlg)

data("cadcm", package = "random.cdisc.data")
data("cadsl", package = "random.cdisc.data")

adsl <- cadsl 

adcm <- cadcm %>% 
  filter(SAFFL == "Y") %>% 
  mutate(CMFL = "Y")

# Create analysis population counts
tbl1 <- freq(adsl,
             rowvar = "SAFFL",
             colvar = "ARM",
             statlist = statlist("n"),
             rowtext = "Analysis Set: Safety Population",
             subset = SAFFL == "Y")

# Create counts (percentages) for patients with any ConMed
tbl2 <- freq(adcm,
             denom_df = adsl,
             rowvar = "CMFL",
             colvar = "ARM",
             statlist = statlist("n (x.x%)"),
             rowtext = "Number of subjects with any concomitant medication",
             subset = CMFL == "Y")

# Create counts (percentages) by CMCLAS and CMDECOD
tbl3 <- nested_freq(adcm,
                    denom_df = adsl,
                    rowvar = "CMCLAS*CMDECOD",
                    colvar = "ARM",
                    statlist = statlist("n (x.x%)"))

# combine analysis results together
tbl <- bind_table(tbl1, tbl2, tbl3) %>% 
  select(-CMCLAS)

# output the analysis results
gentlg(huxme       = tbl,
       format      = "HTML",
       print.hux = FALSE,
       file        = "Table x.x.x.x",
       orientation = "portrait",
       title = "Conmed Example Uses the ‘adcm’ dataset from ‘random.cdisc.data’",
       colheader = c("","A: Drug X","B: Placebo","C: Combination"))

```

### tfrmt 
Rather than starting with an [ADaM](https://www.cdisc.org/standards/foundational/adam), {tfrmt} assumes users will start with an ARD (Analysis Results Dataset), because of this, making this table will be split into two parts, first to make the ARD and second to format the table.
```{r}
resetSession()
library(tidyverse)
library(tfrmt)

## Create ARD 
data("cadcm", package = "random.cdisc.data")
data("cadsl", package = "random.cdisc.data")

big_n <- cadsl |> 
  dplyr::group_by(ARM) |> 
  dplyr::summarize(
    N = dplyr::n_distinct(USUBJID)
  ) 

adcm_with_N <- cadcm |> 
  left_join(big_n, by= "ARM")

overall <- adcm_with_N |> 
  dplyr::group_by(ARM) |>
  dplyr::reframe(
    n_subj = n_distinct(USUBJID),
    pct_subj = n_subj/N
  ) |>
  dplyr::distinct() |>
  dplyr::mutate(CMCLAS = "At Least One Concomittant Med",
         CMDECOD = CMCLAS)

med_lvl <- adcm_with_N |> 
  dplyr::group_by(ARM,CMDECOD, CMCLAS) |>
  dplyr::reframe(
    n_subj = dplyr::n_distinct(USUBJID),
    pct_subj = n_subj/N
  ) |>
  distinct() 

label_N <- big_n |> 
  dplyr::rename(value = N) |> 
  dplyr::mutate(name = "header_n")


cm_ard <- bind_rows(overall, med_lvl) |>
  pivot_longer(ends_with("subj")) |> 
  bind_rows(label_N)

## Format Table 
tfrmt(
  column = ARM,
  group = c("CMCLAS"),
  param = name,
  value = value,
  label = CMDECOD, 
) |>
  # Then we cam combine it with an n percent template 
tfrmt_n_pct(n = "n_subj",
            pct = "pct_subj",
  pct_frmt_when = frmt_when("==1" ~ "", 
                            ">.99" ~ "(>99%)", 
                            "==0" ~ "", 
                            "<.01" ~ "(<1%)", 
                            "TRUE" ~ frmt("(xx.x%)", transform = ~.*100))
  ) |>
  #Finally we are going to add some additional formatting
  tfrmt(
    big_n = big_n_structure("header_n"),
    # Aligning on decimal places and spaces
    col_style_plan = col_style_plan(
      col_style_structure(col = matches("[A-Z]:.*"),
                          align = c(".", " "))
    )
  ) |> 
  print_to_gt(cm_ard)
```

