---
output: html_document
editor_options: 
  chunk_output_type: console
---

## Time to Event Analysis Tables

### Data and models used throughout

```{r, comment=NA}
resetSession()
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(survival)

data("cadaette", package = "random.cdisc.data")
head(cadaette)

adtte <- cadaette %>% 
    dplyr::filter(PARAMCD == "AETTE2", SAFFL == "Y")

```

Cox Proportional Hazard fit:
```{r}
cph <- coxph(Surv(AVAL, CNSR==0) ~ TRT01A + STRATA1, ties = "exact", data = adtte)
```

Kaplan-Meier Model

```{r}
surv_tbl <- as.data.frame(summary(survfit(Surv(AVAL, CNSR==0) ~ TRT01A,
                                          data = adtte, conf.type = "log-log"))$table) %>%
    dplyr::mutate(TRT01A = factor(str_remove(row.names(.), "TRT01A="),
                                  levels = levels(adtte$TRT01A)),
                  ind = FALSE)
.kmState <- currentState()
```

### rtables

```{r, comment=NA}
resetSession(.kmState)
library(rtables)

cnsr_counter <- function(df, .var, .N_col) {
    x <- df[!duplicated(df$USUBJID), .var]
    x <- x[x != "__none__"]
    lapply(table(x), function(xi) rcell(xi*c(1, 1/.N_col), format = "xx (xx.xx%)"))
}
            
a_count_subjs <- function(x, .N_col) {
    in_rows("Subjects with Adverse Events n (%)" = rcell(length(unique(x)) * c(1, 1 / .N_col),
                                                                           format = "xx (xx.xx%)"))
}

a_cph <- function(df, .var, .in_ref_col, .ref_full, full_cox_fit) {
    if(.in_ref_col) {
        ret <- replicate(3, list(rcell(NULL)))
    } else {
        curtrt <- df[[.var]][1]
        coefs <- coef(full_cox_fit)
        sel_pos <- grep(curtrt, names(coefs), fixed = TRUE)
        hrval <- exp(coefs[sel_pos])
        hrvalret <- rcell(hrval, format = "xx.x")
        sdf <- survdiff(Surv(AVAL, CNSR==0) ~ TRT01A + STRATA1,
                        data = rbind(df, .ref_full))
        pval <- (1-pchisq(sdf$chisq, length(sdf$n)-1))/2
        ci_val <- exp(unlist(confint(full_cox_fit)[sel_pos,]))
        ret <- list(rcell(hrval, format = "xx.x"),
                    rcell(ci_val, format = "(xx.x, xx.x)"),
                    rcell(pval, format = "x.xxxx | (<0.0001)"))
    }
    in_rows(.list = ret, .names = c("Hazard ratio",
                                    "95% confidence interval",
                                    "p-value (one-sided stratified log rank)"))
}

a_tte <- function(df, .var,  kp_table) {
    ind <- grep(df[[.var]][1], row.names(kp_table), fixed = TRUE)
    minmax <- range(df[["AVAL"]])

    mm_val_str <- format_value(minmax, format = "xx.x, xx.x")
    rowfn <- list()
    if(all(df$CNSR[df$AVAL == minmax[2]])) {
        mm_val_str <- paste0(mm_val_str, "*")
        rowfn <- "* indicates censoring"
    }
    in_rows(Median = kp_table[ind, "median", drop = TRUE],
            "95% confidence interval" = unlist(kp_table[ind, c("0.95LCL", "0.95UCL")]),
            "Min Max" = mm_val_str,
            .formats = c("xx.xx",
                         "xx.xx - xx.xx",
                         "xx"), .row_footnotes = list(NULL, NULL, rowfn))
}
            

adtte2 <- adtte %>%
    mutate(CNSDTDSC = ifelse(CNSDTDSC == "", "__none__", CNSDTDSC))


lyt <- basic_table(show_colcounts = TRUE,
                   title = "x.x: Safety Data",
                   subtitles = c("x.x.x: Time to First Serious Adverse Event",
                                 "Table x.x.x.x: Safety Endpoint - Safety Analysis Set"),
                   main_footer = c("Serious adverse events are defines as (...). All p-values are exploratory.",
                                   "Hazard ratios are from a stratified Cox model of serious adverse event hazard rate,",
                                   "with terms for treatment groups and strata1. Ties were handled using the exact",
                                   "method. Hazard ratios of Placebo Combination over Drug X are presented, an",
                                   "HR < 1 denotes improvement compared to Drug X."),
                   prov_footer = "Source: ADTTE DDMMYYYY hh:mm; Listing x.xx; SDTM package: DDMMYYYY") %>%
    ## column faceting
    split_cols_by("ARM", ref_group = "A: Drug X") %>%
    ## overall count
    analyze("USUBJID", a_count_subjs, show_labels = "hidden") %>%
    ## Censored Subjects Summary
    analyze("CNSDTDSC", cnsr_counter, var_labels = "Censored Subjects", show_labels = "visible") %>%
    ## CPH analysis
    analyze("ARM", a_cph, extra_args = list(full_cox_fit = cph), show_labels = "hidden") %>%
    ## Time-to-event analysis
    analyze("ARM", a_tte, var_labels = "Time to first adverse event", show_labels = "visible",
            extra_args = list(kp_table = surv_tbl),
            table_names = "kapmeier")

tbl_tte <- build_table(lyt, adtte2)

fnotes_at_path(tbl_tte, c("ma_USUBJID_CNSDTDSC_ARM_kapmeier", "kapmeier")) <- "Product-limit (Kaplan-Meier) estimates."
tbl_tte

```
    
            
    
### Cell Value Derivation for gt of Time to Event Analysis


Our standard TTE table consists of (a derivation of) four main parts:

1. Descriptive stats including the number of subjects with an event, number of subjects censored and censoring reasons
2. Hazard ratio with corresponding 95% CI from a (stratified) Cox model and a p-value from a stratified log rank test
3. Median time to event Kaplan-Meier analysis
4. Number of patients at risk at specified visits from Kaplan-Meier analysis (omitted here).

```{r descr_stats}
resetSession(.kmState)
library(gt)
### Subject Count with events

## surv_tbl calculated above


subj_count <- surv_tbl %>%
  dplyr::mutate(pct = sprintf("%i (%5.1f)", events, 100*events/records),
                label = "Number of subjects with serious adverse event, n (%)") %>%
  dplyr::select(label, TRT01A, pct) %>%
  tidyr::pivot_wider(id_cols = label, names_from = TRT01A, values_from = pct) %>%
  dplyr::mutate(ind = FALSE)

# Number of censored subjects

cnsrd_subj_full <- surv_tbl %>% 
  dplyr::mutate(pct = sprintf("%i (%4.1f)", records-events, 100*(records-events)/records),
                CNSDTDSC = "Number of censored subjects, n (%)") %>% 
  dplyr::select(CNSDTDSC, TRT01A, pct)
  
cnsrd_subj <- adtte %>% 
  dplyr::group_by(TRT01A) %>%
  dplyr::mutate(CNSR = CNSR/n()) %>%
  dplyr::ungroup() %>%
  dplyr::filter(CNSR != 0) %>%
  dplyr::group_by(TRT01A, CNSDTDSC) %>%
  dplyr::summarise(pct = sprintf("%i (%4.1f)", sum(CNSR != 0), 100*sum(CNSR))) %>%
  dplyr::ungroup() %>%
  dplyr::bind_rows(cnsrd_subj_full, .) %>% 
  tidyr::pivot_wider(id_cols = CNSDTDSC, names_from = TRT01A, values_from = pct) %>%
  dplyr::rename(label = CNSDTDSC) %>%
  dplyr::mutate(ind = label != "Number of censored subjects, n (%)") %>% 
  dplyr::arrange(ind)
```

```{r cox_mod}
## cph calculated above
hr <- exp(coef(cph))
ci_hr <- exp(confint(cph))

# Hazard ratio and 95% CI

df_hr <- cbind(ci_hr, hr) %>%
  as.data.frame() %>%
  dplyr::filter(grepl("TRT01A", row.names(.))) %>% 
  dplyr::mutate(TRT01A = factor(str_remove(row.names(.), "TRT01A")),
                ci = sprintf("[%4.1f, %4.1f]", round(!!sym("2.5 %"), 1), round(!!sym("97.5 %"), 1))) %>%
  dplyr::select(TRT01A, hr, ci)

# Log rank p-value

log_rank_test <- purrr::map_df(.x = list(c("A: Drug X", "B: Placebo"),
                                         c("A: Drug X", "C: Combination")),
                               .f = ~{sdf <- survdiff(Surv(AVAL, CNSR==0) ~ TRT01A + STRATA1,
                                                      data = adtte %>% dplyr::filter(TRT01A %in% .x));
                               data.frame(TRT01A = .x[2],
                                          pval = (1-pchisq(sdf$chisq, length(sdf$n)-1))/2)})

df_hr_comp <- merge(df_hr, log_rank_test, by = "TRT01A") %>%
  dplyr::mutate(hr = sprintf("%4.1f", round(hr, 1)),
                pval = ifelse(pval < 0.0001, "<0.0001", sprintf("%6.4f", round(pval, 4)))) %>%
  tidyr::pivot_longer(cols = c(hr, ci, pval), names_to = "label", values_to = "val") %>%
  tidyr::pivot_wider(names_from = TRT01A, values_from = "val") %>%
  dplyr::mutate(label = dplyr::recode(label,
                                      "hr" = "Hazard ratio",
                                      "ci" = "95% confidence interval",
                                      "pval" = "p-value (one-sided stratified log rank)"),
                ind = FALSE)
```

```{r km}
median_survtime <- surv_tbl %>%
  dplyr::mutate(ci = sprintf("[%4.2f, %4.2f]", !!sym("0.95LCL"), !!sym("0.95UCL")),
                median = sprintf("%4.2f", median),
                id = "") %>%
  dplyr::select(TRT01A, id, median, ci) %>%
  tidyr::pivot_longer(cols = c(id, median, ci), names_to = "label", values_to = "val") %>%
  tidyr::pivot_wider(names_from = TRT01A, values_from = val) %>%
  dplyr::mutate(ind = label != "id",
                label = dplyr::recode(label, "median" = "Median (years)",
                                      "ci" = "95% confidence interval",
                                      "id" = "Time to first serious adverse event (a)"))

min_max <- adtte %>%
  dplyr::filter(!(AVAL == 0 & CNSR == 1)) %>% 
  dplyr::group_by(TRT01A) %>%
  dplyr::mutate(max_cnsr = !is.na(AVAL) & AVAL == max(AVAL, na.rm = TRUE) & CNSR == 1) %>%
  dplyr::summarise(min_max = sprintf("%4.2f, %4.2f%s",
                                     min(AVAL, na.rm = TRUE),
                                     max(AVAL, na.rm = TRUE),
                                     ifelse(sum(max_cnsr) > 0, "*", ""))) %>%
  dplyr::ungroup() %>%
  tidyr::pivot_wider(names_from = TRT01A, values_from = min_max) %>%
  dplyr::mutate(label = "Min, Max (b)",
                ind = TRUE)

model_sum <- dplyr::bind_rows(subj_count, cnsrd_subj, df_hr_comp, median_survtime, min_max)
```



### gt

```{r}
header_n <- adtte %>%
  dplyr::group_by(TRT01A) %>%
  dplyr::summarise(TRT = sprintf("%s<br>N=%i (100%%)", unique(TRT01A), n())) %>%
  dplyr::ungroup()


### Begin table creation

gt(model_sum) %>%
  cols_hide(ind) %>%
  tab_header(
    title = "x.x: Safety Data",
    subtitle = html("x.x.x: Time to First Serious Adverse Event<br>Table x.x.x.x: Safety Endpoint - Safety Analysis Set"),
    preheader = c("Protocol: XXXXX", "Cutoff date: DDMMYYYY")
  ) %>%
  tab_source_note("Source: ADTTE DDMMYYYY hh:mm; Listing x.xx; SDTM package: DDMMYYYY") %>%
  opt_align_table_header(align = "left") %>%
  cols_align(align = c("center"),
             columns = c("A: Drug X", "B: Placebo", "C: Combination")) %>%
  cols_align(align = "left",
             columns = "label") %>%
  tab_style(style = cell_text(indent = pct(5)),
            locations = cells_body(columns = 1,
                                   rows = ind == TRUE)) %>%
  sub_missing(columns = everything(), missing_text = "") %>%
  cols_label("label" = "",
             "A: Drug X" = html(header_n$TRT[1]),
             "B: Placebo" = html(header_n$TRT[2]),
             "C: Combination" = html(header_n$TRT[3])) %>%
  tab_footnote(footnote = html("Serious adverse events are defines as (...). All p-values are exploratory.<br>(a) Product-limit (Kaplan-Meier) estimates. <br>(b) Minimum and maximum of event times. * Denotes censoring.<br>Hazard ratios are from a stratified Cox model of serious adverse event hazard rate, with terms for treatment groups and strata1. Ties were handled using the exact method. Hazard ratios of Placebo/ Combination over Drug X are presented, a HR < 1 denotes improvement compared to Drug X.")) %>%
  tab_options(
    table.font.names = "Courier new",
    table.font.size = 9,
    page.orientation = "landscape",
    page.numbering = TRUE,
    page.header.use_tbl_headings = TRUE,
    page.footer.use_tbl_notes = TRUE)
```



### flextable


```{r}
resetSession(.kmState)
ARM_COUNTS <- .ARM_COUNTS # Restore from previous section

library(flextable)
library(tidyverse)

# labels and notes ----
# named vectors that will be used later as 
# labels for the table.
description_labels <- c(SAE = "Number of subjects with serious adverse event", CENSOR_EVENTS = "Number of censored subjects",
  HAZARD_RATIO = "Hazard ratio", TTF_SAE = "Time to first adverse event")
subdescription_labels <- c(subdescription = "",
  ALL = "Censored Subjects", EST = "Estimate", CONFINT = "95% confidence interval",
  PVALUE = "one-sided stratified log rank p-value",
  MEDIAN = "Median", RANGE = "Min Max")
labels <- c(description_labels, subdescription_labels)

additional_notes <- c(
  "Serious adverse events are defines as (...). All p-values are exploratory.", 
  paste("Hazard ratios are from a stratified Cox model", 
        "of serious adverse event hazard rate, with terms for treatment groups and strata1.", 
        "Ties were handled using the exact method. Hazard ratios of Placebo/ Combination over", 
        "Drug X are presented, a HR < 1 denotes improvement compared to Drug X."))
titles <- c(
  "x.x: Safety Data",
  "x.x.x: Time to First Serious Adverse Event",
  "Table x.x.x.x: Safety Endpoint - Safety Analysis Set")

# Data need to be prepared in a single data.frame, 
# a function is written to make the code simplier to organize.
#' @noRd
#' @description create a data.frame made of 
#' several sub data.frame, each for a specific 
#' calculation. Final result has two specific columns 
#' 'description' and 'subdescription' that will 
#' contain various labels or group labels.
prepare_tte_df <- function(adtte, cph, surv_tbl) {
  ## Number of subjects with serious adverse event ----
  subj_count <- surv_tbl |>
    select(TRT01A, n = events, records) |>
    mutate(pct = n / records, description = "SAE", subdescription = "ALL")

  ## Number of censored subjects ----
  subj_censored <- subj_count |>
    mutate(
      n = records - n, pct = 1 - pct, description = "CENSOR_EVENTS",
      subdescription = "ALL"
    )

  ## Number of censored subjects - detail per CNSDTDSC ----
  subj_censored_detail <- adtte |>
    filter(CNSR > 0) |>
    count(TRT01A, CNSDTDSC) |>
    left_join(subj_count |> select(TRT01A, records), by = "TRT01A") |>
    mutate(pct = n / records, description = "CENSOR_EVENTS") |>
    rename(subdescription = "CNSDTDSC")

  ## Hazard ratio and 95% CI ----
  coefs_hr <- cbind(exp(confint(cph)), y = exp(coef(cph))) |> as.data.frame() |>
    rownames_to_column(var = "TRT01A") |>
    filter(grepl("TRT01A", TRT01A)) |>
    mutate(TRT01A = str_replace(TRT01A, "TRT01A", ""))

  hr_dat <- bind_rows(
    select(coefs_hr, TRT01A, LCL = `2.5 %`, UCL = `97.5 %`) |>
      mutate(description = "HAZARD_RATIO", subdescription = "CONFINT"),
    select(coefs_hr, TRT01A, y) |>
      mutate(description = "HAZARD_RATIO", subdescription = "EST")
  )

  ## p-value (one-sided stratified log rank)	 ----
  pval_dat <- purrr::map_df(
    .x = list(c("A: Drug X", "B: Placebo"), c("A: Drug X", "C: Combination")),
    .f = ~ {
      sdf <- survdiff(Surv(AVAL, CNSR == 0) ~ TRT01A + STRATA1, data = adtte |> 
                        dplyr::filter(TRT01A %in% .x))
      data.frame(TRT01A = .x[2], description = "HAZARD_RATIO", subdescription = "PVALUE",
        y = (1 - pchisq(sdf$chisq, length(sdf$n) - 1)) / 2)
    }
  )

  ## Time to first serious adverse event ----
  tfe_dat <- bind_rows(
    surv_tbl |>
      select(TRT01A, LCL = `0.95LCL`, UCL = `0.95UCL`) |>
      mutate(description = "TTF_SAE", subdescription = "CONFINT"),
    surv_tbl |>
      select(TRT01A, y = median) |>
      mutate(description = "TTF_SAE", subdescription = "MEDIAN"),
    adtte |>
      filter(!(AVAL == 0 & CNSR == 1)) |>
      group_by(TRT01A) |>
      mutate(max_cnsr = !is.na(AVAL) & AVAL == max(AVAL, na.rm = TRUE) & CNSR == 1) |>
      summarise(
        min = min(AVAL, na.rm = TRUE),
        max = max(AVAL, na.rm = TRUE),
        is_censored = sum(max_cnsr) > 0,
        description = "TTF_SAE", subdescription = "RANGE"
      )
  )
  
  cnsdtdsc_levels <- unique(adtte$CNSDTDSC) |> sort() |> setdiff("")

  dat <- bind_rows(subj_count, subj_censored, subj_censored_detail, hr_dat, pval_dat, tfe_dat) |> 
    as_tibble() |>
    ## factor for easy ordering ----
    mutate(
      description = factor(description, levels = c("SAE", "CENSOR_EVENTS", "HAZARD_RATIO", "TTF_SAE")),
      subdescription = factor(subdescription,
      levels = c("ALL", cnsdtdsc_levels, "NUSUBJID", "MEDIAN", "EST", "CONFINT", "PVALUE", "RANGE")))

  dat
}
dat <- prepare_tte_df(adtte, cph, surv_tbl)

# Now data.frame is ready to be formatted with `fmt_time_to_event()`. 
# The function handle each special cases to be covered so that 
# the four main parts are formatted as expected.

library(glue)
library(dplyr)
# format fun ----
fmt_time_to_event <- function(description, subdescription, y, cts, pcts, LCL, UCL, min, max, is_censored) {
  case_when(
    !is.na(cts) & !is.na(pcts) ~ glue('{sprintf("%.0f", cts)} ({sprintf("%.02f", pcts*100)}%)'),
    !is.na(y) & !subdescription %in% "PVALUE" ~ glue('{sprintf("%.02f", y)}'),
    !is.na(y) & subdescription %in% "PVALUE" ~ glue("{scales::label_pvalue()(y)}"),
    !is.na(LCL) & !is.na(UCL) ~ glue('[{sprintf("%.02f", LCL)}, {sprintf("%.02f", UCL)}]'),
    !is.na(LCL) & !is.na(UCL) ~ glue('[{sprintf("%.02f", LCL)}, {sprintf("%.02f", UCL)}]'),
    !is.na(min) & !is.na(max) & is_censored ~ glue('{sprintf("%.02f", min)}, {sprintf("%.02f", max)}*'),
    !is.na(min) & !is.na(max) & !is_censored ~ glue('{sprintf("%.02f", min)}, {sprintf("%.02f", max)}'),
    TRUE ~ ""
  )
}

# use `tabulator()` and `as_flextable()` together to create the table. 

tab <- tabulator(dat,
  rows = c("description", "subdescription"), columns = "TRT01A",
  content_cell = as_paragraph(
    fmt_time_to_event(
      description = description, subdescription = subdescription,
      y = y, cts = n, pcts = pct, LCL = LCL, UCL = UCL,
      min = min, max = max, is_censored = is_censored))
)
ft <- as_flextable(x = tab, spread_first_col = TRUE, separate_with = "description", sep_w = 0) |>
  # footnotes
  footnote(
    j = 1, i = ~ description %in% c("CENSOR_EVENTS", "SAE"),
    ref_symbols = " (1)", value = as_paragraph(" Number of subjects (% of subjects)")
  ) |>
  footnote(
    j = 1, i = ~ description %in% "TTF_SAE" & is.na(subdescription), ref_symbols = " (2)",
    value = as_paragraph(" Product-limit (Kaplan-Meier) estimates.")
  ) |>
  footnote(
    j = 1, i = ~ subdescription %in% "RANGE", ref_symbols = " (3)",
    value = as_paragraph(" * indicates censoring")
  ) |>
  add_footer_lines(values = additional_notes) |>
  # labels
  labelizor(j = "subdescription", labels = labels, part = "all") |>
  # indentation
  prepend_chunks(i = ~ is.na(description), j = "subdescription", as_chunk("\t")) |>
  align(i = ~ !is.na(description), align = "left") |>
  autofit() |>
  add_header_lines(values = titles) |>
  hline_top(part = "header", border = fp_border_default(width = 0)) |>
  hline(part = "header", i = 1:2, border = fp_border_default(width = 0))

for (ARM_COD in names(ARM_COUNTS)) {
  ft <- append_chunks(x = ft, part = "header", i = 4,
    j = tabulator_colnames(tab, columns = "content_cell", TRT01A %in% !!ARM_COD),
    as_chunk(ARM_COUNTS[ARM_COD], formatter = fmt_header_n))
}
ft
```

### tables

```{r}
resetSession(.kmState)

library(tables)
table_options(doCSS = TRUE)

ex_adae <- formatters::ex_adae

subject_counts <- table(adsl$ARM)

countpercentid <- function(num, ARM) {
  n <- length(unique(num))
  sprintf("%d (%.2f%%)", 
          length(unique(num)), 
          100*n/subject_counts[ARM[1]])
}

valuepercent <- function(x, ARM) {
  sprintf("%d (%.2f%%)", x, 100*x/subject_counts[ARM] )
}

blanks <- function(x) ""

count <- function(x) sprintf("(N=%d)", length(x))

hazardratio <- function(ARM) {
  entry <- paste0("TRT01A", ARM)
  coef <- coef(cph)
  if (entry %in% names(coef)) sprintf("%.1f", exp(coef[entry]))
  else ""
}

hazardratioconfint <- function(ARM) {
  entry <- paste0("TRT01A", ARM)
  confint <- confint(cph)
  if (entry %in% rownames(confint)) {
    confint <- as.numeric(confint[entry,])
    sprintf("(%.1f, %.1f)", exp(confint[1]), exp(confint[2]))
  } else ""
}

hazardpvalue <- function(ARM) {
  if (ARM == "A: Drug X") ""
  else {
    twogroups <- c("A: Drug X", ARM)
    sdf <- survdiff(Surv(AVAL, CNSR==0) ~ TRT01A + STRATA1,
                   data = adtte, subset = TRT01A %in% twogroups)
    pval <- (1-pchisq(sdf$chisq, length(sdf$n)-1))/2
    sprintf("%.4f", pval)
  }
}

Median <- function(ARM) {
  vals <- subset(surv_tbl, TRT01A == ARM)
  sprintf("%.2f", vals$median)
}

minmaxevent <- function(ARM) {
  vals <- subset(adtte, TRT01A == ARM)
  sprintf("%.2f, %.2f", min(vals$AVAL), max(vals$AVAL))
}

eventCI <- function(ARM) {
  vals <- subset(surv_tbl, TRT01A == ARM)
  sprintf("[%.2f, %.2f]", vals$`0.95LCL`, vals$`0.95UCL`)
}

heading <- tabular(Heading("")*count ~ 
                   Heading()*ARM, 
                   data = adsl)

part1 <- tabular( Heading("Subjects with serious adverse events")*
                    events*Heading()*
                    valuepercent*Arguments(ARM = TRT01A) ~ 
                  Heading()*TRT01A, 
                  data = surv_tbl )

part2 <- tabular( Heading("Number of censored subjects")*
                    blanks +
                  Factor(CNSDTDSC, "")*
                    Heading()*countpercentid*Arguments(ARM = TRT01A)*
                    Heading()*USUBJID ~
                  Heading()*TRT01A, 
                  data = subset(adtte, nchar(CNSDTDSC) > 0))

part3 <- tabular( ( Heading("Hazard ratio")*hazardratio +
                    Heading("95% confidence interval")*hazardratioconfint +
                    Heading("p-value (one-sided stratified log rank)")*hazardpvalue +
                    Heading("Time to first serious adverse event")*blanks +
                    Heading("Median (years)")*Median +
                    Heading("95% confidence interval")*eventCI +
                    Heading("Min, Max")*minmaxevent)*
                   Heading()*as.character(TRT01A)  ~
                  Heading()*TRT01A,
                  data = surv_tbl)

rbind(heading, part1, part2, part3)
```

### tidytlg

```{r}
resetSession(.kmState)
library(dplyr)
library(tidytlg)
library(broom)
library(stringr)

# Create analysis population counts
tbl1 <- freq(adtte,
             rowvar = "SAFFL",
             colvar = "TRT01A",
             statlist = statlist("n"),
             rowtext = "Analysis Set: Safety Population",
             subset = SAFFL == "Y")

# Create counts (percentages) for subjects with SAE
tbl2 <- freq(adtte,
             rowvar = "CNSR",
             colvar = "TRT01A",
             statlist = statlist("n (x.x%)"),
             rowtext = "Number of subjects with serious adverse events, n(%)",
             subset = CNSR == 0)

# Create counts (percentages) for subjects with SAE
tbl3a <- freq(adtte,
             rowvar = "CNSR",
             colvar = "TRT01A",
             statlist = statlist("n (x.x%)"),
             rowtext = "Number of censored subjects, n(%)",
             subset = CNSR == 1)

tbl3b <- freq(adtte,
             rowvar = "CNSDTDSC",
             colvar = "TRT01A",
             statlist = statlist("n (x.x%)"),
             subset = CNSR == 1)

tbl3 <- bind_rows(tbl3a, tbl3b)

# CoxPH model
coxmod <- tidy(cph, exponentiate = TRUE, conf.int = TRUE, conf.level = 0.95) %>% 
  filter(str_detect(term, "TRT01A")) %>% 
  mutate(term = str_remove(term, "TRT01A"))

tbl4a <- coxmod %>% 
  mutate(hr = roundSAS(estimate, digits = 2, as_char = TRUE)) %>% 
  select(term, hr) %>% 
  pivot_wider(names_from = "term", values_from = "hr") %>% 
  mutate(label = "Hazard ratio",
         row_type = "HEADER")

tbl4b <- coxmod %>% 
  mutate(across(c(conf.low, conf.high), ~roundSAS(.x, digits = 2)),
         ci = paste0("(", conf.low, ", ", conf.high, ")")) %>% 
  select(term, ci) %>% 
  pivot_wider(names_from = "term", values_from = "ci") %>% 
  mutate(label = "95% Confidence Interval",
         row_type = "VALUE")

tbl4 <- bind_rows(tbl4a, tbl4b) %>% 
  mutate(group_level = 0)

# Logrank test
log_rank_test <- purrr::map_df(.x = list(c("A: Drug X", "B: Placebo"),
                                         c("A: Drug X", "C: Combination")),
                               .f = ~{sdf <- survdiff(Surv(AVAL, CNSR==0) ~ TRT01A + STRATA1,
                                                      data = adtte %>% dplyr::filter(TRT01A %in% .x));
                               data.frame(TRT01A = .x[2],
                                          pval = (1-pchisq(sdf$chisq, length(sdf$n)-1))/2)})

tbl5 <- log_rank_test %>% 
  mutate(pval = roundSAS(pval, digits = 3, as_char = TRUE)) %>% 
  pivot_wider(names_from = "TRT01A", values_from = "pval") %>% 
  mutate(label = "p-value (one-sided stratified log rank)",
         row_type = "HEADER",
         group_level = 0)

# surv time stats
tbl6a <- surv_tbl %>% 
  mutate(median = roundSAS(median, digits = 2, as_char = TRUE)) %>% 
  select(TRT01A, median) %>% 
  pivot_wider(names_from = "TRT01A", values_from = "median") %>% 
  mutate(label = "Median (years)",
         row_type = "VALUE") %>% 
  add_row(label = "Time to first serious adverse event (a)", row_type = "HEADER", .before = 1)

tbl6b <- surv_tbl %>% 
  mutate(across(c(`0.95LCL`, `0.95UCL`), ~roundSAS(.x, digits = 2, as_char = TRUE)),
         ci = paste0("(", `0.95LCL`, ", ", `0.95UCL`, ")")) %>% 
  select(TRT01A, ci) %>% 
  pivot_wider(names_from = "TRT01A", values_from = "ci") %>% 
  mutate(label = "95% Confidence Interval",
         row_type = "VALUE")

tbl6c <- adtte %>% 
  filter(!(AVAL == 0 & CNSR == 1)) %>% 
  group_by(TRT01A) %>% 
  mutate(max_cnsr = !is.na(AVAL) & AVAL == max(AVAL, na.rm = TRUE) & CNSR == 1) %>% 
  summarise(
        min = min(AVAL, na.rm = TRUE),
        max = max(AVAL, na.rm = TRUE),
        is_censored = sum(max_cnsr) > 0) %>% 
  mutate(across(c(min, max), ~roundSAS(.x, digits = 2, as_char = TRUE)),
         min_max = ifelse(is_censored, paste0("(", min, ", ", max, "*)"), 
                          paste0("(", min, ", ", max, ")"))) %>% 
  select(TRT01A, min_max) %>% 
  pivot_wider(names_from = "TRT01A", values_from = "min_max") %>% 
  mutate(label = "Min - Max (b)",
         row_type = "VALUE")


tbl6 <- bind_rows(tbl6a, tbl6b, tbl6c) %>% 
  mutate(group_level = 0)

# combine analysis results together
tbl <- bind_table(tbl1, tbl2, tbl3, tbl4, tbl5, tbl6)

# output the analysis results
gentlg(huxme       = tbl,
       format      = "HTML",
       print.hux = FALSE,
       file        = "Table x.x.x.x",
       orientation = "portrait",
       title = "Time to First Serious Adverse Event",
       colheader = c("","A: Drug X","B: Placebo","C: Combination"))  


```


