---
output: html_document
editor_options: 
  chunk_output_type: console
---

## Disposition

### rtables

```{r, comment=NA}
resetSession()

library(dplyr)
library(rtables)

data("cadsl", package = "random.cdisc.data")
adsl <- cadsl |>
  select(USUBJID, TRT01A, EOSSTT, DCSREAS, DTHCAUS)

top_afun <- function(x, .N_col) {
    in_rows(Completed = rcell(sum(x=="COMPLETED") * c(1, 1/.N_col), format = c("xx (xx.x%)")),
            Ongoing = rcell(sum(x=="ONGOING") * c(1, 1/.N_col), format = c("xx (xx.x%)")))
}

count_pct_afun <- function(x, .N_col) {
    tbl <- table(x)
    lst <- lapply(tbl, function(xi) rcell(xi * c(1, 1/.N_col), format = c("xx (xx.x%)")))
    in_rows(.list = lst, .names = names(tbl))
}
            
lyt <- basic_table(show_colcounts = TRUE) |>
  split_cols_by("TRT01A") |>
  analyze("EOSSTT", top_afun) |>
  split_rows_by("EOSSTT", split_fun = keep_split_levels("DISCONTINUED")) |>
  analyze("DCSREAS", count_pct_afun) |>
  split_rows_by("DCSREAS", split_fun = keep_split_levels("DEATH")) |>
  analyze("DTHCAUS", count_pct_afun)

build_table(lyt, adsl)

```


### tern (+ rtables)

*tern* makes slightly different formatting choices (2 decimals for
the percents of completed and ongoing study counts, and not displaying
the percent when a cell count is 0), but we can see the table
structure and cell values are the same.

```{r}
library(tern)
lyt <- basic_table(show_colcounts = TRUE) |>
  split_cols_by("TRT01A") |>
  count_values("EOSSTT",
               values = "COMPLETED",
               table_names = "Completed",
               .labels = c(count_fraction = "Completed Study")) |>
  count_values("EOSSTT",
               values = "ONGOING",
               table_names = "Ongoing",
               .labels = c(count_fraction = "Ongoing Study")) |>
  split_rows_by("EOSSTT",
                split_fun = keep_split_levels("DISCONTINUED")) |>
  summarize_vars("DCSREAS",
                 .stats = "count_fraction",
                 denom = "N_col") |>
  split_rows_by("DCSREAS",
                split_fun = keep_split_levels("DEATH")) |>
  summarize_vars("DTHCAUS",
                 .stats = "count_fraction",
                 denom = "N_col")

build_table(lyt = lyt, df = adsl)
```


### gt

```{r}
resetSession()

library(tidyverse)
library(gt)

adsl_tot <- cadsl |> 
  dplyr::summarize(
    NTOT = dplyr::n(),
    NTOTLBL = sprintf("%s  \n(N=%i)", unique(TRT01A), dplyr::n()),
    .by = TRT01A
  )

header_n <- as.list(adsl_tot$NTOTLBL)
names(header_n) <- paste0("n_", adsl_tot$TRT01A)

disp_df <- merge(cadsl, adsl_tot, by = "TRT01A") |> 
  dplyr::mutate(
    EOSSTT = factor(EOSSTT, levels = c("COMPLETED", "ONGOING", "DISCONTINUED"))
  )

disc_status <- disp_df |> 
  dplyr::filter(EOSSTT != "DISCONTINUED") |> 
  dplyr::summarize(
    n = dplyr::n(),
    pct = dplyr::n()/mean(NTOT),
    .by = c(TRT01A, EOSSTT)
  ) |> 
  tidyr::pivot_wider(id_cols = EOSSTT, names_from = TRT01A, values_from = c(n, pct)) |> 
  dplyr::arrange(EOSSTT)

disc_reason <- disp_df |> 
  dplyr::filter(EOSSTT == "DISCONTINUED") |> 
  dplyr::summarize(
    n = dplyr::n(),
    pct = dplyr::n()/mean(NTOT),
    .by = c(TRT01A, EOSSTT, DCSREAS)
  ) |> 
  tidyr::pivot_wider(id_cols = c(EOSSTT, DCSREAS), names_from = TRT01A, values_from = c(n, pct)) |> 
  dplyr::arrange(EOSSTT, DCSREAS)

disc_death <- disp_df |> 
  dplyr::filter(DCSREAS == "DEATH") |> 
  dplyr::mutate(
    EOSSTT = "DEATH",
    DCSREAS = DTHCAUS
    ) |> 
  dplyr::summarize(
    n = dplyr::n(),
    pct = dplyr::n()/mean(NTOT),
    .by = c(TRT01A, EOSSTT, DCSREAS)
  ) |> 
  tidyr::pivot_wider(id_cols = c(EOSSTT, DCSREAS), names_from = TRT01A, values_from = c(n, pct)) |> 
  dplyr::arrange(EOSSTT, DCSREAS)

gt_disp <- dplyr::bind_rows(disc_status, disc_reason, disc_death) 

gt_disp |> 
  gt(rowname_col = "DCSREAS") |> 
  tab_row_group(
    label = "Discontinued",
    rows = EOSSTT == "DISCONTINUED"
  ) |> 
  tab_row_group(
    label = "Death",
    rows = EOSSTT == "DEATH"
  ) |> 
  row_group_order(
    groups = c(NA, "Discontinued", "Death") 
  ) |> 
  fmt_integer(
    columns = starts_with("n_")
  ) |> 
  fmt_percent(
    columns = starts_with("pct_"),
    decimals = 2
  ) |> 
  cols_merge_n_pct(col_n = "n_A: Drug X", col_pct = "pct_A: Drug X") |> 
  cols_merge_n_pct(col_n = "n_B: Placebo", col_pct = "pct_B: Placebo") |> 
  cols_merge_n_pct(col_n = "n_C: Combination", col_pct = "pct_C: Combination") |> 
  cols_merge(
    columns = c("DCSREAS", "EOSSTT"),
    rows = EOSSTT %in% c("COMPLETED", "ONGOING"),
    pattern = "<<{1}>><<{2}>>"
  ) |> 
  sub_missing(
    columns = starts_with("n_"),
    missing_text = "0"
  ) |> 
  text_transform(
    locations = list(cells_body(), cells_stub()),
    fn = stringr::str_to_title
  ) |> 
  cols_align(
    align = "left",
    columns = "DCSREAS"
  ) |> 
  cols_align(
    align = "center",
    columns = starts_with("n_")
  ) |>  
  cols_label(
    .list = header_n,
    .fn = md
  )  |> 
  tab_stub_indent(
    rows = 3:16,
    indent = 5
  ) |> 
  cols_width(
    1 ~ px(200)
  )
```


### flextable

```{r}
resetSession()

library(survival)
library(tidyverse)
library(flextable)
library(glue)

adsl <- cadsl |>
  select(USUBJID, TRT01A, EOSSTT, DCSREAS)

# data parts calculations
part_header <- adsl |> count(TRT01A, name = "n_part")

part_completed <- adsl |> filter(EOSSTT %in% "COMPLETED") |> 
  mutate(DCSREAS = "") |>
  count(TRT01A, EOSSTT, DCSREAS)

part_ongoing <- adsl |> filter(EOSSTT %in% "ONGOING") |> 
  mutate(DCSREAS = "") |>
  count(TRT01A, EOSSTT, DCSREAS)

part_discontinued <- adsl |> 
  filter(EOSSTT %in% "DISCONTINUED") |> 
  count(TRT01A, EOSSTT, DCSREAS)

part_death <- cadsl |> 
  filter(EOSSTT %in% "DISCONTINUED", DCSREAS %in% "DEATH") |> 
  count(TRT01A, EOSSTT, DTHCAUS) |> 
  mutate(DTHCAUS = paste0("\t", DTHCAUS)) |> 
  rename(DCSREAS = DTHCAUS)

DCSREAS_LEV <- c(
  "", "ADVERSE EVENT", "DEATH", 
  part_death$DCSREAS, levels(part_discontinued$DCSREAS)) |> 
  unique()
EOSSTT_LEV <- c("COMPLETED", "ONGOING", "DISCONTINUED")

dat <- bind_rows(
  part_completed, 
  part_ongoing, 
  part_discontinued, 
  part_death) |> 
  inner_join(part_header, by = "TRT01A") |> 
  mutate(percent = n / n_part, n_part = NULL,
         DCSREAS = factor(DCSREAS, levels = DCSREAS_LEV),
         EOSSTT = factor(EOSSTT, levels = EOSSTT_LEV)
  )

# Now the flextable creation with help of `tabulator()`. 

tab <- tabulator(
  dat,
  rows = c("EOSSTT", "DCSREAS"),
  columns = "TRT01A",
  `content_cell` = as_paragraph(fmt_n_percent(n, percent))
)
ft <- as_flextable(tab, spread_first_col = TRUE, 
                   columns_alignment = "center" )

TRT_COUNTS <- setNames(part_header$n_part, part_header$TRT01A)
for (TRT_COD in names(TRT_COUNTS)) {
  ft <- append_chunks(x = ft, part = "header", i = 1,
                      j = tabulator_colnames(tab, columns = "content_cell", TRT01A %in% !!TRT_COD),
                      as_chunk(TRT_COUNTS[TRT_COD], formatter = function(n) sprintf("\n(N=%.0f)", n)))
}

ft <- labelizor(ft, j = "DCSREAS", part = "all", labels = function(x) tools::toTitleCase(tolower(x))) |> 
  labelizor(labels = c(Dcsreas = ""), j = "DCSREAS", part = "header") |> 
  align(i = ~!is.na(EOSSTT) | seq_along(EOSSTT) == 1, j = 1, align = "left") |> 
  prepend_chunks(i = ~is.na(EOSSTT), j = "DCSREAS", as_chunk("\t")) |> 
  autofit()
ft
```

### tables

```{r}
resetSession()

adsl <- cadsl

# Change the labels to title case

levels(adsl$EOSSTT)  <- tools::toTitleCase(tolower(levels(adsl$EOSSTT)))
levels(adsl$DCSREAS) <- tools::toTitleCase(tolower(levels(adsl$DCSREAS)))
levels(adsl$DTHCAUS) <- tools::toTitleCase(tolower(levels(adsl$DTHCAUS)))

library(tables)

subject_counts <- table(adsl$ARM)

countpercentid <- function(num, ARM) {
  n <- length(unique(num))
  if (n == 0) pct <- 0
  else        pct <- 100*n/subject_counts[ARM[1]]
  sprintf("%d (%.2f%%)", 
          length(unique(num)), 
          pct)
}

count <- function(x) sprintf("(N=%d)", length(x))

heading <- tabular(Heading("")*1*Heading("")*count  ~
             Heading()*TRT01A, data = adsl)

part1 <- tabular( Heading("")*EOSSTT*DropEmpty(which = "row")*
                    Heading("")*1*
                    Heading()*countpercentid*Arguments(ARM = TRT01A)*
                    Heading()*USUBJID ~
                  Heading()*TRT01A, 
                  data = subset(adsl, EOSSTT != "Discontinued"))

part2 <- tabular( Heading("")*EOSSTT*
                    Heading("")*DCSREAS*DropEmpty(which = "row")*
                    Heading()*countpercentid*Arguments(ARM = TRT01A)*
                    Heading()*USUBJID ~
                  Heading()*TRT01A, 
                  data = subset(adsl, EOSSTT == "Discontinued" &
                                      DCSREAS != "Death"))

part3 <- tabular( Heading("")*DCSREAS*
                    Heading("")*DTHCAUS*DropEmpty(which = "row")*
                    Heading()*countpercentid*Arguments(ARM = TRT01A)*
                    Heading()*USUBJID ~
                  Heading()*TRT01A, 
                  data = subset(adsl, EOSSTT == "Discontinued" &
                                      DCSREAS == "Death"))

useGroupLabels(rbind(heading, part1, part2, part3), 
               indent = "&emsp;")
```

### tidytlg

```{r}
resetSession()
library(dplyr)
library(tidytlg)

data("cadsl", package = "random.cdisc.data")

adsl <- cadsl %>% 
  mutate(COMPFL = case_when(EOSSTT == "COMPLETED" ~ "Y",
                            TRUE ~ "N"))

disc <- adsl %>% 
  filter(EOSSTT == "DISCONTINUED")

dth <- adsl %>% 
  filter(DTHFL == "Y")

# Create analysis population counts
tbl1 <- freq(adsl,
             rowvar = "SAFFL",
             colvar = "ARM",
             statlist = statlist("n"),
             rowtext = "Analysis Set: Safety Population",
             subset = SAFFL == "Y")

# Create counts (percentages) for completed patients
tbl2 <- freq(adsl,
             rowvar = "COMPFL",
             colvar = "ARM",
             statlist = statlist("n (x.x%)"),
             rowtext = "Completed",
             subset = COMPFL == "Y")

# Create counts (percentages) for discontinued reasons
tbl3 <- freq(disc,
             denom_df = adsl,
             rowvar = "DCSREAS",
             colvar = "ARM",
             statlist = statlist("n (x.x%)"),
             row_header = "Discontinued")

# Create counts (percentages) for death reasons
tbl4 <- freq(dth,
             denom_df = adsl,
             rowvar = "DTHCAUS",
             colvar = "ARM",
             statlist = statlist("n (x.x%)"),
             row_header = "Death Cause")

# combine analysis results together
tbl <- bind_table(tbl1, tbl2, tbl3, tbl4)

# output the analysis results
gentlg(huxme       = tbl,
       format      = "HTML",
       print.hux = FALSE,
       file        = "Table x.x.x.x",
       orientation = "portrait",
       title = "Study Disposition Summary",
       colheader = c("","A: Drug X","B: Placebo","C: Combination"))
```


### tfrmt
Rather than starting with an [ADaM](https://www.cdisc.org/standards/foundational/adam), *tfrmt* assumes users will start with an ARD (Analysis Results Dataset), because of this, making this table will be split into two parts, first to make the ARD and second to format the table. 
```{r}
resetSession()

library(tidyverse)
library(tfrmt)

data("cadsl", package = "random.cdisc.data")


big_n <- cadsl |> 
  dplyr::group_by(ARM) |> 
  dplyr::summarize(
    N = dplyr::n_distinct(USUBJID)
  ) 

disp_summary <- cadsl |>
  dplyr::left_join(big_n, by = "ARM") |> 
  dplyr::group_by(ARM, EOSSTT, DCSREAS, DTHCAUS) |> 
  dplyr::reframe(
     n_subj = n_distinct(USUBJID),
     pct_subj = n_subj/N
  ) |> 
  dplyr::distinct() |> 
  tidyr::pivot_longer(ends_with("subj")) |> 
  dplyr::mutate(
    DCSREAS = if_else(is.na(DCSREAS), EOSSTT, DCSREAS),
    DTHCAUS = if_else(is.na(DTHCAUS), DCSREAS, DTHCAUS),
    EOSSTT = forcats::fct_relevel(EOSSTT, 
                                  "ONGOING", "COMPLETED", "DISCONTINUED")
  ) %>% 
  dplyr::arrange(EOSSTT, DCSREAS, DTHCAUS)

label_N <- big_n |> 
  dplyr::rename(value = N) |> 
  dplyr::mutate(name = "header_n")


disp_ard <- disp_summary |>
  bind_rows(label_N)

## Format Table 
tfrmt(
  column = ARM,
  group = c("EOSSTT", "DCSREAS"),
  param = name,
  value = value,
  label = DTHCAUS
) |>
  # Then we cam combine it with an n percent template 
tfrmt_n_pct(n = "n_subj",
            pct = "pct_subj",
  pct_frmt_when = frmt_when("==1" ~ "", 
                            ">.99" ~ "(>99%)", 
                            "==0" ~ "", 
                            "<.01" ~ "(<1%)", 
                            "TRUE" ~ frmt("(xx.x%)", transform = ~.*100)) 
  ) |>
  #Finally we are going to add some additional formatting
  tfrmt(
    big_n = big_n_structure("header_n"),
    # Aligning on decimal places and spaces
    col_style_plan = col_style_plan(
      col_style_structure(col = matches("[A-Z]:.*"),
                          align = c(".", " "))
    )
  )|> 
  print_to_gt(disp_ard)
```

