---
output: html_document
editor_options: 
  chunk_output_type: console
---

## Time to Event Analysis Tables

### Data and models used throughout

```{r, comment=NA}
resetSession()
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(survival)

data("cadaette", package = "random.cdisc.data")
head(cadaette)

adtte <- cadaette %>% 
    dplyr::filter(PARAMCD == "AETTE2", SAFFL == "Y")

```

Cox Proportional Hazard fit:
```{r}
cph <- coxph(Surv(AVAL, CNSR==0) ~ TRT01A + STRATA1, ties = "exact", data = adtte)
```

Kaplan-Meier Model

```{r}
surv_tbl <- as.data.frame(summary(survfit(Surv(AVAL, CNSR==0) ~ TRT01A,
                                          data = adtte, conf.type = "log-log"))$table) %>%
    dplyr::mutate(TRT01A = factor(str_remove(row.names(.), "TRT01A="),
                                  levels = levels(adtte$TRT01A)),
                  ind = FALSE)
mn_footer_txt <- c("Serious adverse events are defined as (...). All p-values are exploratory.",
                   "Hazard ratios are from a stratified Cox model of serious adverse event hazard rate,",
                   "with terms for treatment groups and strata1. Ties were handled using the exact",
                   "method. Hazard ratios of Placebo Combination over Drug X are presented, an",
                   "HR < 1 denotes improvement compared to Drug X.")

stitle_txt <- c("x.x.x: Time to First Serious Adverse Event",
                "Table x.x.x.x: Safety Endpoint - Safety Analysis Set")
.kmState <- currentState()
```

### rtables

```{r, comment=NA}
resetSession(.kmState)
library(rtables)

## this will be properly exported in the next release of rtables
RefFootnote <- rtables:::RefFootnote

cnsr_counter <- function(df, .var, .N_col) {
    x <- df[!duplicated(df$USUBJID), .var]
    x <- x[x != "__none__"]
    lapply(table(x), function(xi) rcell(xi*c(1, 1/.N_col), format = "xx (xx.xx%)"))
}
            
a_count_subjs <- function(x, .N_col) {
    in_rows("Subjects with Adverse Events n (%)" = rcell(length(unique(x)) * c(1, 1 / .N_col),
                                                                           format = "xx (xx.xx%)"))
}

a_cph <- function(df, .var, .in_ref_col, .ref_full, full_cox_fit) {
    if(.in_ref_col) {
        ret <- replicate(3, list(rcell(NULL)))
    } else {
        curtrt <- df[[.var]][1]
        coefs <- coef(full_cox_fit)
        sel_pos <- grep(curtrt, names(coefs), fixed = TRUE)
        hrval <- exp(coefs[sel_pos])
        hrvalret <- rcell(hrval, format = "xx.x")
        sdf <- survdiff(Surv(AVAL, CNSR==0) ~ TRT01A + STRATA1,
                        data = rbind(df, .ref_full))
        pval <- (1-pchisq(sdf$chisq, length(sdf$n)-1))/2
        ci_val <- exp(unlist(confint(full_cox_fit)[sel_pos,]))
        ret <- list(rcell(hrval, format = "xx.x"),
                    rcell(ci_val, format = "(xx.x, xx.x)"),
                    rcell(pval, format = "x.xxxx | (<0.0001)"))
    }
    in_rows(.list = ret, .names = c("Hazard ratio",
                                    "95% confidence interval",
                                    "p-value (one-sided stratified log rank)"))
}

a_tte <- function(df, .var,  kp_table) {
    ind <- grep(df[[.var]][1], row.names(kp_table), fixed = TRUE)
    minmax <- range(df[["AVAL"]])

    mm_val_str <- format_value(minmax, format = "xx.x, xx.x")
    rowfn <- list()
    in_rows(Median = kp_table[ind, "median", drop = TRUE],
            "95% confidence interval" = unlist(kp_table[ind, c("0.95LCL", "0.95UCL")]),
            "Min Max" = mm_val_str,
            .formats = c("xx.xx",
                         "xx.xx - xx.xx",
                         "xx"), .cell_footnotes = list(NULL, NULL, list(RefFootnote("Denotes censoring", index = 0L, symbol = "*"))))
}
            

adtte2 <- adtte |>
    mutate(CNSDTDSC = ifelse(CNSDTDSC == "", "__none__", CNSDTDSC))

lyt <- basic_table(show_colcounts = TRUE,
                   title = "x.x: Safety Data",
                   subtitles = stitle_txt,
                                 "Table x.x.x.x: Safety Endpoint - Safety Analysis Set"),
                   main_footer = mn_footer_txt,
                   prov_footer = "Source: ADTTE DDMMYYYY hh:mm; Listing x.xx; SDTM package: DDMMYYYY") |>
    split_cols_by("ARM", ref_group = "A: Drug X") |>
    analyze("USUBJID", a_count_subjs, show_labels = "hidden") |>
    analyze("CNSDTDSC", cnsr_counter, var_labels = "Censored Subjects", show_labels = "visible") |>
    analyze("ARM", a_cph, extra_args = list(full_cox_fit = cph), show_labels = "hidden") |>
    analyze("ARM", a_tte, var_labels = "Time to first adverse event", show_labels = "visible",
            extra_args = list(kp_table = surv_tbl),
            table_names = "kapmeier")

tbl_tte <- build_table(lyt, adtte2)

fnotes_at_path(tbl_tte, c("ma_USUBJID_CNSDTDSC_ARM_kapmeier", "kapmeier")) <- "Product-limit (Kaplan-Meier) estimates."
tbl_tte

```

### tern (+rtables)

*tern* encapsulates the specific statistical choices used by Roche. In
particulate, its implementation of the Cox pairwise analysis does not
implement the one-tailed p-value strategy used in the rest of this
chapter.

We will first showcaes the pure *tern* solution, which has different
p-values for this reason, and then implement a hybrid *tern* +
explicit *rtables* solution which fully recreates the exact table
generated by other systems.

```{r}
resetSession(.kmState)
library(tern)

## this will be properly exported in the next release of rtables
RefFootnote <- rtables:::RefFootnote

adtte3 <- adtte
adtte3$is_event <- adtte$CNSR == 0
adtte3$CNSDTDSC[adtte$CNSDTDSC == ""] <- NA



lyt1 <- basic_table(show_colcounts = TRUE,
                   title = "x.x: Safety Data",
                   subtitles = stitle_txt,
                   main_footer = mn_footer_txt,
                   prov_footer = "Source: ADTTE DDMMYYYY hh:mm; Listing x.xx; SDTM package: DDMMYYYY") |>
  split_cols_by("TRT01A", ref_group = "A: Drug X") |>
  count_values(
    "STUDYID",
    values = "AB12345",
    .stats = "count_fraction",
    .labels = c(count_fraction = "Subjects with Adverse Events n (%)")
  ) |>
  count_occurrences(
    "CNSDTDSC",
    var_labels = "Censored Subjects",
    show_labels = "visible",
    .formats = c(count_fraction = "xx.xx (xx.xx%)")
  ) |>
  coxph_pairwise(
    vars = "AVAL",
    is_event = "is_event",
    control = control_coxph(pval_method = "log-rank", ties = "exact"),
    strat = "STRATA1",
    .stats = c("hr", "hr_ci", "pvalue"),
    .formats = c(hr = "xx.x", hr_ci = "(xx.x, xx.x)", pvalue = "xx.xxxx"),
    .labels = c(hr = "Hazard ratio", hr_ci = "95% confidence interval", pvalue = "p-value (stratified log rank)"),
    show_labels = "hidden",
    table_names = "coxph"
  ) |>
  surv_time(
    vars = "AVAL",
    is_event = "is_event",
    control = control_surv_time(conf_type = "log-log"),
    .stats = c("median", "median_ci", "range"),
    .formats = c(median = "xx.xx", median_ci = "xx.xx - xx.xx", range = "xx.x, xx.x"),
    .labels = c(median_ci = "95% confidence interval", range = "Min Max"),
    .indent_mods = c(median_ci = 0L),
    var_labels = "Time to first adverse event"
  )

tbl_tte_tern <- build_table(lyt = lyt1, df = adtte3)

fnotes_at_path(tbl_tte_tern, c("ma_STUDYID_CNSDTDSC_coxph_AVAL", "AVAL")) <- "Product-limit (Kaplan-Meier) estimates."

fnote <- RefFootnote("Censored.", index = 0L, symbol = "^")

for(pth in col_paths(tbl_tte_tern)) {
    fnotes_at_path(tbl_tte_tern,
                   rowpath = c("ma_STUDYID_CNSDTDSC_coxph_AVAL", "AVAL", "range"),
                   colpath = pth) <- fnote
}
tbl_tte_tern
```

We now create the hybrid table where we utilize a custom analysis
function to recreate the one-sided p-values while using *tern* for the
rest of the table structure.

```{r}


a_cph <- function(df, .var, .in_ref_col, .ref_full, full_cox_fit) {
  if(.in_ref_col) {
    ret <- replicate(3, list(rcell(NULL)))
  } else {
    curtrt <- df[[.var]][1]
    coefs <- coef(full_cox_fit)
    sel_pos <- grep(curtrt, names(coefs), fixed = TRUE)
    hrval <- exp(coefs[sel_pos])
    hrvalret <- rcell(hrval, format = "xx.x")
    sdf <- survival::survdiff(Surv(AVAL, CNSR==0) ~ TRT01A + STRATA1,
                              data = rbind(df, .ref_full))
    pval <- (1-pchisq(sdf$chisq, length(sdf$n)-1))/2
    ci_val <- exp(unlist(confint(full_cox_fit)[sel_pos,]))
    ret <- list(rcell(hrval, format = "xx.x"),
                rcell(ci_val, format = "(xx.x, xx.x)"),
                rcell(pval, format = "x.xxxx | (<0.0001)"))
  }
  in_rows(.list = ret, .names = c("Hazard ratio",
                                  "95% confidence interval",
                                  "p-value (one-sided stratified log rank)"))
}

lyt2 <- basic_table(show_colcounts = TRUE,
                   title = "x.x: Safety Data",
                   subtitles = stitle_txt,
                   main_footer = mn_footer_txt,
                   prov_footer = "Source: ADTTE DDMMYYYY hh:mm; Listing x.xx; SDTM package: DDMMYYYY") |>
  split_cols_by("TRT01A", ref_group = "A: Drug X") |>
  count_values(
    "STUDYID",
    values = "AB12345",
    .stats = "count_fraction",
    .labels = c(count_fraction = "Subjects with Adverse Events n (%)")
  ) |>
  count_occurrences(
    "CNSDTDSC",
    var_labels = "Censored Subjects",
    show_labels = "visible",
    .formats = c(count_fraction = "xx.xx (xx.xx%)")
  ) |>
  analyze("ARM", a_cph, extra_args = list(full_cox_fit = cph), show_labels = "hidden") |>
  surv_time(
    vars = "AVAL",
    is_event = "is_event",
    control = control_surv_time(conf_type = "log-log"),
    .stats = c("median", "median_ci", "range"),
    .formats = c(median = "xx.xx", median_ci = "xx.xx - xx.xx", range = "xx.x, xx.x"),
    .labels = c(median_ci = "95% confidence interval", range = "Min Max"),
    .indent_mods = c(median_ci = 0L),
    var_labels = "Time to first adverse event"
  )

tbl_tte_tern2 <- build_table(lyt = lyt2, df = adtte3)

fnotes_at_path(tbl_tte_tern2, c("ma_STUDYID_CNSDTDSC_ARM_AVAL", "AVAL")) <- "Product-limit (Kaplan-Meier) estimates."

fnote <- RefFootnote("Denotes censoring.", index = 0L, symbol = "*")

for(pth in col_paths(tbl_tte_tern2)) {
    fnotes_at_path(tbl_tte_tern2,
                   rowpath = c("ma_STUDYID_CNSDTDSC_ARM_AVAL", "AVAL", "range"),
                   colpath = pth) <- fnote
}
tbl_tte_tern2

```

### Cell Value Derivation for gt of Time to Event Analysis


Our standard TTE table consists of (a derivation of) four main parts:

1. Descriptive stats including the number of subjects with an event, number of subjects censored and censoring reasons
2. Hazard ratio with corresponding 95% CI from a (stratified) Cox model and a p-value from a stratified log rank test
3. Median time to event Kaplan-Meier analysis
4. Number of patients at risk at specified visits from Kaplan-Meier analysis (omitted here).

```{r descr_stats}
resetSession(.kmState)
library(gt)
### Subject Count with events

## surv_tbl calculated above


subj_count <- surv_tbl  |> 
  dplyr::mutate(pct = sprintf("%i (%5.1f)", events, 100*events/records),
                label = "Number of subjects with serious adverse event, n (%)") |>
  dplyr::select(label, TRT01A, pct) |>
  tidyr::pivot_wider(id_cols = label, names_from = TRT01A, values_from = pct) |>
  dplyr::mutate(ind = FALSE)

# Number of censored subjects

cnsrd_subj_full <- surv_tbl |>
  dplyr::mutate(pct = sprintf("%i (%4.1f)", records-events, 100*(records-events)/records),
                CNSDTDSC = "Number of censored subjects, n (%)") |> 
  dplyr::select(CNSDTDSC, TRT01A, pct)
  
cnsrd_subj <- adtte |> 
  dplyr::group_by(TRT01A) |>
  dplyr::mutate(CNSR = CNSR/n()) |>
  dplyr::ungroup() |>
  dplyr::filter(CNSR != 0) |>
  dplyr::group_by(TRT01A, CNSDTDSC) |>
  dplyr::summarise(pct = sprintf("%i (%4.1f)", sum(CNSR != 0), 100*sum(CNSR)), .groups = "drop") |>
  dplyr::bind_rows(cnsrd_subj_full) |> 
  tidyr::pivot_wider(id_cols = CNSDTDSC, names_from = TRT01A, values_from = pct) |>
  dplyr::rename(label = CNSDTDSC) |>
  dplyr::mutate(ind = label != "Number of censored subjects, n (%)") |>
  dplyr::arrange(ind)
```

```{r cox_mod}
## cph calculated above
hr <- exp(coef(cph))
ci_hr <- exp(confint(cph))

# Hazard ratio and 95% CI

df_hr <- cbind(ci_hr, hr) |>
  as.data.frame() |>
  (\(data) dplyr::filter(data, grepl("TRT01A", row.names(data))))() |> 
  (\(data) dplyr::mutate(
    data, 
    TRT01A = factor(stringr::str_remove(row.names(data), "TRT01A")),
    ci = sprintf("[%4.1f, %4.1f]", round(!!sym("2.5 %"), 1), round(!!sym("97.5 %"), 1))
    ))() |> 
  dplyr::select(TRT01A, hr, ci)

# Log rank p-value

log_rank_test <- purrr::map_df(.x = list(c("A: Drug X", "B: Placebo"),
                                         c("A: Drug X", "C: Combination")),
                               .f = ~{sdf <- survdiff(Surv(AVAL, CNSR==0) ~ TRT01A + STRATA1,
                                                      data = adtte |> dplyr::filter(TRT01A %in% .x));
                               data.frame(TRT01A = .x[2],
                                          pval = (1-pchisq(sdf$chisq, length(sdf$n)-1))/2)})

df_hr_comp <- merge(df_hr, log_rank_test, by = "TRT01A") |>
  dplyr::mutate(hr = sprintf("%4.1f", round(hr, 1)),
                pval = ifelse(pval < 0.0001, "<0.0001", sprintf("%6.4f", round(pval, 4)))) |>
  tidyr::pivot_longer(cols = c(hr, ci, pval), names_to = "label", values_to = "val") |>
  tidyr::pivot_wider(names_from = TRT01A, values_from = "val") |>
  dplyr::mutate(label = dplyr::recode(label,
                                      "hr" = "Hazard ratio",
                                      "ci" = "95% confidence interval",
                                      "pval" = "p-value (one-sided stratified log rank)"),
                ind = FALSE)
```

```{r km}
median_survtime <- surv_tbl |>
  dplyr::mutate(ci = sprintf("[%4.2f, %4.2f]", !!sym("0.95LCL"), !!sym("0.95UCL")),
                median = sprintf("%4.2f", median),
                id = "") |>
  dplyr::select(TRT01A, id, median, ci) |>
  tidyr::pivot_longer(cols = c(id, median, ci), names_to = "label", values_to = "val") |>
  tidyr::pivot_wider(names_from = TRT01A, values_from = val) |>
  dplyr::mutate(ind = label != "id",
                label = dplyr::recode(label, "median" = "Median (years)",
                                      "ci" = "95% confidence interval",
                                      "id" = "Time to first serious adverse event (a)"))

min_max <- adtte |>
  dplyr::filter(!(AVAL == 0 & CNSR == 1)) |> 
  dplyr::group_by(TRT01A) |>
  dplyr::mutate(max_cnsr = !is.na(AVAL) & AVAL == max(AVAL, na.rm = TRUE) & CNSR == 1) |>
  dplyr::summarize(
    min_max = sprintf("%4.2f, %4.2f%s", min(AVAL, na.rm = TRUE), max(AVAL, na.rm = TRUE), ifelse(sum(max_cnsr) > 0, "*", "")),
    .groups = "drop"
    ) |>
  tidyr::pivot_wider(names_from = TRT01A, values_from = min_max) |>
  dplyr::mutate(label = "Min, Max (b)",
                ind = TRUE)

model_sum <- dplyr::bind_rows(subj_count, cnsrd_subj, df_hr_comp, median_survtime, min_max)
```



### gt

```{r}
header_n <- adtte |> 
  dplyr::group_by(TRT01A) |>
  dplyr::summarise(N = dplyr::n(), .groups = "drop") |> 
  dplyr::transmute(TRT = sprintf("%s  \n  N=%i (100%%)", TRT01A, N)) |> 
  dplyr::group_split(TRT)


### Begin table creation

gt(model_sum) |> 
  cols_hide(ind) |>
  tab_header(
    title = "x.x: Safety Data",
    subtitle = md("x.x.x: Time to First Serious Adverse Event  \n  Table x.x.x.x: Safety Endpoint - Safety Analysis Set"),
    preheader = c("Protocol: XXXXX", "Cutoff date: DDMMYYYY")
  ) |> 
  tab_source_note("Source: ADTTE DDMMYYYY hh:mm; Listing x.xx; SDTM package: DDMMYYYY") |>
  opt_align_table_header(align = "left") |>
  cols_align(align = c("center"),
             columns = c("A: Drug X", "B: Placebo", "C: Combination")) |>
  cols_align(align = "left",
             columns = "label") |>
  tab_style(style = cell_text(indent = pct(5)),
            locations = cells_body(columns = 1,
                                   rows = ind == TRUE)) |>
  sub_missing(columns = everything(), missing_text = "") |>
  cols_label("label" = "",
             "A: Drug X" = md(header_n[[1]]),
             "B: Placebo" = md(header_n[[2]]),
             "C: Combination" = md(header_n[[3]])) |> 
  tab_footnote(footnote = md("Serious adverse events are defines as (...). All p-values are exploratory.  \n  (a) Product-limit (Kaplan-Meier) estimates.   \n  (b) Minimum and maximum of event times. * Denotes censoring.  \n  Hazard ratios are from a stratified Cox model of serious adverse event hazard rate, with terms for treatment groups and strata1. Ties were handled using the exact method. Hazard ratios of Placebo/ Combination over Drug X are presented, a HR < 1 denotes improvement compared to Drug X.")) |> 
  tab_options(
    table.font.names = "Courier new",
    table.font.size = 9,
    page.orientation = "landscape",
    page.numbering = TRUE,
    page.header.use_tbl_headings = TRUE,
    page.footer.use_tbl_notes = TRUE)
```





### tables

```{r}
resetSession(.kmState)

library(tables)
table_options(doCSS = TRUE)

ex_adae <- formatters::ex_adae

subject_counts <- table(adsl$ARM)

countpercentid <- function(num, ARM) {
  n <- length(unique(num))
  if (n == 0) pct <- 0
  else        pct <- 100*n/subject_counts[ARM[1]]
  sprintf("%d (%.2f%%)", 
          length(unique(num)), 
          pct)
}

valuepercent <- function(x, ARM) {
  sprintf("%d (%.2f%%)", x, 100*x/subject_counts[ARM] )
}

blanks <- function(x) ""

count <- function(x) sprintf("(N=%d)", length(x))

hazardratio <- function(ARM) {
  entry <- paste0("TRT01A", ARM)
  coef <- coef(cph)
  if (entry %in% names(coef)) sprintf("%.1f", exp(coef[entry]))
  else ""
}

hazardratioconfint <- function(ARM) {
  entry <- paste0("TRT01A", ARM)
  confint <- confint(cph)
  if (entry %in% rownames(confint)) {
    confint <- as.numeric(confint[entry,])
    sprintf("(%.1f, %.1f)", exp(confint[1]), exp(confint[2]))
  } else ""
}

hazardpvalue <- function(ARM) {
  if (ARM == "A: Drug X") ""
  else {
    twogroups <- c("A: Drug X", ARM)
    sdf <- survdiff(Surv(AVAL, CNSR==0) ~ TRT01A + STRATA1,
                   data = adtte, subset = TRT01A %in% twogroups)
    pval <- (1-pchisq(sdf$chisq, length(sdf$n)-1))/2
    sprintf("%.4f", pval)
  }
}

Median <- function(ARM) {
  vals <- subset(surv_tbl, TRT01A == ARM)
  sprintf("%.2f", vals$median)
}

minmaxevent <- function(ARM) {
  vals <- subset(adtte, TRT01A == ARM)
  sprintf("%.2f, %.2f", min(vals$AVAL), max(vals$AVAL))
}

eventCI <- function(ARM) {
  vals <- subset(surv_tbl, TRT01A == ARM)
  sprintf("[%.2f, %.2f]", vals$`0.95LCL`, vals$`0.95UCL`)
}

heading <- tabular(Heading("")*1*Heading("")*count ~ 
                   Heading()*ARM, 
                   data = adsl)

part1 <- tabular( Heading("Subjects with serious adverse events")*1*Heading("")*
                    events*Heading()*
                    valuepercent*Arguments(ARM = TRT01A) ~ 
                  Heading()*TRT01A, 
                  data = surv_tbl )

part2 <- tabular( Heading("Number of censored subjects")*1*Factor(CNSDTDSC, "")*
                Heading()*countpercentid*Arguments(ARM = TRT01A)*
                    Heading()*USUBJID ~
                  Heading()*TRT01A, 
                  data = subset(adtte, nchar(CNSDTDSC) > 0))

part3 <- tabular( ( Heading("Hazard ratio")*1*Heading("")*hazardratio +
                    Heading("95% confidence interval")*1*Heading("")*hazardratioconfint +
                    Heading("p-value (one-sided stratified log rank)")*1*Heading("")*hazardpvalue +
                    Heading("Time to first serious adverse event")*1*(
                      Heading("Median (years)")*Median +
                      Heading("95% confidence interval")*eventCI +
                      Heading("Min, Max")*minmaxevent))*
                    Heading()*as.character(TRT01A)  ~
                  Heading()*TRT01A,
                  data = surv_tbl)

useGroupLabels(rbind(heading, part1, part2, part3),
               indent = "&emsp;")
```

### flextable

This is a situation where the code required to create a flextable directly
becomes too long or complex. In such case, it is more convenient to leverage
existing functions from other packages to generate a tabular object and then
convert it to a flextable using the `as_flextable()` method. Here we reuse 
the `tables` objects created in the previous section.

```{r}
library(flextable)
rbind(part1, part2, part3) |> 
  as_flextable(spread_first_col = TRUE, add_tab = TRUE) |>
  align(j = 1, part = "all", align = "left") |> 
  padding(padding = 4, part = "all") |> 
  add_header_row(
    values = c("", fmt_header_n(subject_counts, newline = FALSE)),
    top = FALSE
  ) |> 
  hline(i = 1, part = "header", border = fp_border_default(width = 0))
```


### tidytlg

```{r}
resetSession(.kmState)
library(dplyr)
library(tidytlg)
library(broom)
library(stringr)

# Create analysis population counts
tbl1 <- freq(adtte,
             rowvar = "SAFFL",
             colvar = "TRT01A",
             statlist = statlist("n"),
             rowtext = "Analysis Set: Safety Population",
             subset = SAFFL == "Y")

# Create counts (percentages) for subjects with SAE
tbl2 <- freq(adtte,
             rowvar = "CNSR",
             colvar = "TRT01A",
             statlist = statlist("n (x.x%)"),
             rowtext = "Number of subjects with serious adverse events, n(%)",
             subset = CNSR == 0)

# Create counts (percentages) for subjects with SAE
tbl3a <- freq(adtte,
             rowvar = "CNSR",
             colvar = "TRT01A",
             statlist = statlist("n (x.x%)"),
             rowtext = "Number of censored subjects, n(%)",
             subset = CNSR == 1)

tbl3b <- freq(adtte,
             rowvar = "CNSDTDSC",
             colvar = "TRT01A",
             statlist = statlist("n (x.x%)"),
             subset = CNSR == 1)

tbl3 <- bind_rows(tbl3a, tbl3b)

# CoxPH model
coxmod <- tidy(cph, exponentiate = TRUE, conf.int = TRUE, conf.level = 0.95) %>% 
  filter(str_detect(term, "TRT01A")) %>% 
  mutate(term = str_remove(term, "TRT01A"))

tbl4a <- coxmod %>% 
  mutate(hr = roundSAS(estimate, digits = 2, as_char = TRUE)) %>% 
  select(term, hr) %>% 
  pivot_wider(names_from = "term", values_from = "hr") %>% 
  mutate(label = "Hazard ratio",
         row_type = "HEADER")

tbl4b <- coxmod %>% 
  mutate(across(c(conf.low, conf.high), ~roundSAS(.x, digits = 2)),
         ci = paste0("(", conf.low, ", ", conf.high, ")")) %>% 
  select(term, ci) %>% 
  pivot_wider(names_from = "term", values_from = "ci") %>% 
  mutate(label = "95% Confidence Interval",
         row_type = "VALUE")

tbl4 <- bind_rows(tbl4a, tbl4b) %>% 
  mutate(group_level = 0)

# Logrank test
log_rank_test <- purrr::map_df(.x = list(c("A: Drug X", "B: Placebo"),
                                         c("A: Drug X", "C: Combination")),
                               .f = ~{sdf <- survdiff(Surv(AVAL, CNSR==0) ~ TRT01A + STRATA1,
                                                      data = adtte %>% dplyr::filter(TRT01A %in% .x));
                               data.frame(TRT01A = .x[2],
                                          pval = (1-pchisq(sdf$chisq, length(sdf$n)-1))/2)})

tbl5 <- log_rank_test %>% 
  mutate(pval = roundSAS(pval, digits = 3, as_char = TRUE)) %>% 
  pivot_wider(names_from = "TRT01A", values_from = "pval") %>% 
  mutate(label = "p-value (one-sided stratified log rank)",
         row_type = "HEADER",
         group_level = 0)

# surv time stats
tbl6a <- surv_tbl %>% 
  mutate(median = roundSAS(median, digits = 2, as_char = TRUE)) %>% 
  select(TRT01A, median) %>% 
  pivot_wider(names_from = "TRT01A", values_from = "median") %>% 
  mutate(label = "Median (years)",
         row_type = "VALUE") %>% 
  add_row(label = "Time to first serious adverse event (1)", row_type = "HEADER", .before = 1)

tbl6b <- surv_tbl %>% 
  mutate(across(c(`0.95LCL`, `0.95UCL`), ~roundSAS(.x, digits = 2, as_char = TRUE)),
         ci = paste0("(", `0.95LCL`, ", ", `0.95UCL`, ")")) %>% 
  select(TRT01A, ci) %>% 
  pivot_wider(names_from = "TRT01A", values_from = "ci") %>% 
  mutate(label = "95% Confidence Interval",
         row_type = "VALUE")

tbl6c <- adtte %>% 
  filter(!(AVAL == 0 & CNSR == 1)) %>% 
  group_by(TRT01A) %>% 
  mutate(max_cnsr = !is.na(AVAL) & AVAL == max(AVAL, na.rm = TRUE) & CNSR == 1) %>% 
  summarise(
        min = min(AVAL, na.rm = TRUE),
        max = max(AVAL, na.rm = TRUE),
        is_censored = sum(max_cnsr) > 0) %>% 
  mutate(across(c(min, max), ~roundSAS(.x, digits = 2, as_char = TRUE)),
         min_max = ifelse(is_censored, paste0("(", min, ", ", max, "*)"), 
                          paste0("(", min, ", ", max, ")"))) %>% 
  select(TRT01A, min_max) %>% 
  pivot_wider(names_from = "TRT01A", values_from = "min_max") %>% 
  mutate(label = "Min - Max (2)",
         row_type = "VALUE")


tbl6 <- bind_rows(tbl6a, tbl6b, tbl6c) %>% 
  mutate(group_level = 0)

# combine analysis results together
tbl <- bind_table(tbl1, tbl2, tbl3, tbl4, tbl5, tbl6)

# output the analysis results
gentlg(huxme       = tbl,
       format      = "HTML",
       print.hux = FALSE,
       file        = "Table x.x.x.x",
       orientation = "portrait",
       title = "Time to First Serious Adverse Event",
       footers = c("(1) Product-limit (Kaplan-Meier) estimates.",
                   "(2) * indicates censoring",
                   "Serious adverse events are defines as (...). All p-values are exploratory.",
                   "Hazard ratios are from a stratified Cox model of serious adverse event hazard rate,
with terms for treatment groups and strata1. Ties were handled using the exact
method. Hazard ratios of Placebo Combination over Drug X are presented, an
HR < 1 denotes improvement compared to Drug X.",
"Source: ADTTE DDMMYYYY hh:mm; Listing x.xx; SDTM package: DDMMYYYY"),
       colheader = c("","A: Drug X","B: Placebo","C: Combination"))  


```

### tfrmt 

This first code chunk cleans up the data from the models to prepare it for going into a table.
```{r}
library(tidyverse)
library(broom)

big_n <- surv_tbl |> 
  dplyr::select(N = n.max, TRT01A)

# Number of subjects with a serious AE 
sae_n <- surv_tbl |> # Calculated above 
  dplyr::mutate(pct = events/n.max, 
                group = "Number of subjects with serious adverse event, n (%)",
                label = "Number of subjects with serious adverse event, n (%)",
                ord1 = 1 ) |>
  dplyr::select(TRT01A, n = events, pct, group, label, ord1) |> 
  tidyr::pivot_longer(c("n", "pct"))

# Count the number of censored subjects
adtte_with_N <- adtte |>
  dplyr::left_join(big_n, by = "TRT01A")

cnsr_subjs <- adtte_with_N |> 
  dplyr::filter(CNSR == "1") 

tot_cnsr_subj <- cnsr_subjs |> 
  dplyr::group_by(TRT01A) |> 
  dplyr::reframe(
    n = n_distinct(USUBJID),
    pct = n/N
  ) |> 
  dplyr::distinct() |> 
  tidyr::pivot_longer(c("n", "pct")) |> 
  dplyr::mutate(
    group = "Number of censored subjects, n (%)",
    label = "Number of censored subjects, n (%)",
    ord1 = 2
  )

sub_cnsr_subj <- cnsr_subjs |> 
  dplyr::group_by(TRT01A, CNSDTDSC) |> 
  dplyr::reframe(
    n = n_distinct(USUBJID),
    pct = n/N
  ) |> 
  dplyr::distinct() |> 
  tidyr::pivot_longer(c("n", "pct")) |> 
  dplyr::mutate(
    group = "Number of censored subjects, n (%)",
    ord1 = 2
  ) |> 
  dplyr::rename(label = CNSDTDSC)

# Information from the CPH model 
hzr <- broom::tidy(cph, conf.int = TRUE) |> 
  mutate(across(c("estimate", "conf.low", "conf.high"), exp)) |> 
  dplyr::filter(stringr::str_detect(term, "TRT01A")) |>
  dplyr::select(term, estimate, conf.low, conf.high) |> 
  tidyr::pivot_longer(c("estimate", "conf.low", "conf.high")) |> 
  dplyr::mutate(group = "Hazard ratio",
                label = case_when(
                  name == "estimate" ~ "Hazard ratio",
                  TRUE ~ "95% confidence interval"
                ),
                TRT01A = case_when(
                  stringr::str_detect(term, "Placebo") ~ "B: Placebo",
                  stringr::str_detect(term, "Combination") ~ "C: Combination"
                ),
                ord1 = 3) |> 
  dplyr::select(-term)

# Get one-sided p-value from survival model
p_vals <- list(c("A: Drug X", "B: Placebo"), c("A: Drug X", "C: Combination")) |> 
  map_dfr(function(comparison){
    survdiff(Surv(AVAL, CNSR == 0) ~ TRT01A + STRATA1, data = adtte |> 
                        dplyr::filter(TRT01A %in% comparison)) |> 
    broom::glance() |> 
      dplyr::mutate(TRT01A = comparison[2])
  }) |> 
  dplyr::select(value = p.value, TRT01A) |> 
  dplyr::mutate(
    name = "p.value",
    group = "p-value (one-sided stratified log rank)",
    label = "p-value (one-sided stratified log rank)",
    ord1 = 5
  )


# Time to event from model 
time_to_event <- surv_tbl |> 
  dplyr::select(TRT01A, median, LCL = `0.95LCL`, UCL=`0.95UCL`) |> 
  tidyr::pivot_longer(c("median", "LCL", "UCL")) |> 
  dplyr::mutate(
    group = "Time to first serious adverse event",
    label = case_when(
      name == "median" ~ "Median (years)",
      TRUE ~ "95% confidence interval"
    ),
    ord1 = 6
  )

range <- adtte |> 
  dplyr::group_by(TRT01A) |> 
  dplyr::summarise(
    min = min(AVAL),
    max = max(AVAL)
  ) |> 
  dplyr::mutate(group = "Time to first serious adverse event",
                label = "Min, Max",
                ord1 = 6)|> 
  tidyr::pivot_longer(c("min", "max"))


model_ard <- bind_rows(
  sae_n,
  tot_cnsr_subj,
  sub_cnsr_subj,
  hzr,
  p_vals,
  time_to_event,
  range
)
```

We now format this information into a table. 
```{r}
library(tfrmt)
tfrmt(
  column = TRT01A,
  group = "group",
  label = "label",
  param = "name", 
  value = "value", 
  sorting_cols = "ord1",
  body_plan = body_plan(
    frmt_structure(group_val = ".default", label_val = ".default", 
                   frmt_combine("{n} ({pct})",
                                                                                n = frmt("xx"),
                                                                                pct = frmt("xx%", transform = ~.*100))
                   
    ),
    frmt_structure(group_val = "Hazard ratio", label_val = ".default",
                   frmt_combine("[{conf.low}, {conf.high}]",
                                frmt("x.x"))),
    frmt_structure(group_val = ".default", label_val = "Hazard ratio", frmt("x.x")),
    frmt_structure(group_val = ".default", label_val = "p-value (one-sided stratified log rank)", frmt("x.xxxx")),
    frmt_structure(group_val = ".default", label_val = "Median (years)", frmt("x.xx")),
    frmt_structure(group_val = "Time to first serious adverse event", label_val = "95% confidence interval",
                   frmt_combine("[{LCL}, {UCL}]",
                                frmt("x.xx"))),
    frmt_structure(group_val = ".default", label_val = "Min, Max",
                   frmt_combine("[{min}, {max}*]",
                                frmt("x.xx")))
    
  ), 
  col_plan = col_plan(-ord1), 
  footnote_plan = footnote_plan(
    footnote_structure("Serious adverse events are defines as (...). All p-values are exploratory.
                       Hazard ratios are from a stratified Cox model of serious adverse event hazard rate, with terms for treatment groups and strata1. Ties were handled using the exact method. Hazard ratios of Placebo/ Combination over Drug X are presented, a HR < 1 denotes improvement compared to Drug X."),
    footnote_structure(group_val = "Time to first serious adverse event",
                       "Product-limit (Kaplan-Meier) estimates"), 
    footnote_structure(group_val = "Time to first serious adverse event", 
                       label_val = "Min, Max", "Minimum and maximum of event times. * Denotes censoring")
  )
) |> 
  print_to_gt(model_ard)
```
