---
output: html_document
editor_options: 
  chunk_output_type: console
---

## Event Tables

We will use the `ex_adae` data included within the `formatters` package.

```{r, comment=NA}
head(formatters::ex_adae)
```

### Adverse Events

 

#### `rtables` Only

Adverse Events by ID

```{r, comment=NA}
resetSession()
library(rtables)

s_events_patients <- function(x, labelstr, .N_col) {
  in_rows(
    "Patients with at least one event" =
      rcell(length(unique(x)) * c(1, 1 / .N_col), format = "xx (xx.xx%)"),

    "Total number of events" = rcell(length(x), format = "xx")
  )
}

table_count_per_id <- function(df, termvar = "AEDECOD", idvar = "USUBJID") {

  x <- df[[termvar]]
  id <- df[[idvar]]

  counts <- table(x[!duplicated(id)])

  in_rows(
    .list = as.vector(counts),
    .labels = names(counts)
  )
}

lyt <- basic_table(show_colcounts = TRUE) %>%
    split_cols_by("ARM") %>%
    analyze("USUBJID", afun = s_events_patients) %>%
    split_rows_by("AEBODSYS", child_labels = "visible", 
                  split_fun = trim_levels_in_group("AEDECOD")) %>%
    summarize_row_groups("USUBJID", cfun = s_events_patients) %>%
    analyze("AEDECOD", table_count_per_id, show_labels = "hidden", indent_mod = -1)

build_table(lyt, ex_adae, alt_counts_df = ex_adsl)
```

#### flextable

```{r message=FALSE}
resetSession()
ex_adsl <- formatters::ex_adsl
ex_adae <- formatters::ex_adae
library(flextable)
library(tidyr)
library(dplyr)
library(forcats)
ae_tabulator <- function(ADSL, ADAE){
  
  
  subject_cts <- group_by(ADSL, ARM) |> 
    summarise(n_subject = n())
  
  all_event_total_cts <- group_by(ADAE, ARM) |> 
    summarise(n_usub = n_distinct(USUBJID), n_events = n())
  
  aebodsys_cts <- group_by(ADAE, ARM, AEBODSYS) |> 
    summarise(n_usub = n_distinct(USUBJID), n_events = n(),
              .groups = "drop")
  
  aedecod_cts <- group_by(ADAE, ARM, AEBODSYS, AEDECOD) |> 
    summarise(n_usub = n_distinct(USUBJID), n_events = n(),
              .groups = "drop")
  
  x <- bind_rows(all_event_total_cts, aebodsys_cts, aedecod_cts) |> 
    mutate(
      AEBODSYS = fct_expand(AEBODSYS, "ALL1") |> fct_relevel("ALL1"),
      AEDECOD = fct_expand(AEDECOD, "ALL2") |> fct_relevel("ALL2")
    ) |> 
    left_join(subject_cts, by = "ARM") |> 
    replace_na(replace = list(AEBODSYS = "ALL1", AEDECOD = "ALL2")) |>
    arrange(AEBODSYS, AEDECOD) |> 
    mutate(pct = n_usub / n_subject)
  attr(x, "ARM_COUNTS") <- setNames(subject_cts$n_subject, subject_cts$ARM)
  x
  
}
```

Labels for `flextable::labelizor()`:

```{r}
AEDECOD_LABELS <- c(
  ALL1 = "Patients with at least one event", 
  ALL2 = "Patients with at least one event", 
  `GASTROINTESTINAL DISORDERS` = "Gastrointestinal Disorders", 
  `NERVOUS SYSTEM DISORDERS` = "Nervous System Disorders", 
  `VASCULAR DISORDERS` = "Vascular Disorders", 
  `MUSCULOSKELETAL AND CONNECTIVE TISSUE DISORDERS` = "Musculoskeletal and Connective Tissue Disorders", 
  `NAUSEA (INTERMITTENT)` = "Nausea (Intermittent)", 
  HEADACHE = "Headache", 
  `FAECES SOFT` = "Faeces Soft", 
  HYPOTENSION = "Hypotension", 
  `ABDOMINAL FULLNESS DUE TO GAS` = "Abdominal Fullness Due to Gas", 
  `ORTHOSTATIC HYPOTENSION` = "Orthostatic Hypotension", 
  `BACK PAIN` = "Back Pain", 
  WEAKNESS = "Weakness", 
  `GINGIVAL BLEEDING` = "Gingival Bleeding", 
  `ABDOMINAL DISCOMFORT` = "Abdominal Discomfort", 
  DIARRHEA = "Diarrhea"
)
HEADER_LABELS <- c(AEDECOD = "dictionary-derived term\n\tbody system or organ class",
  nevents = "Total number\nof events", 
  n = "n\n(% of subjects)",
  "ARM A" = "Treatment A",
  "ARM B" = "Treatment B"
)
```

`tabulator()` call:

```{r}
dat <- ae_tabulator(ADSL = ex_adsl, ADAE = ex_adae)
ARM_COUNTS <- attr(dat, "ARM_COUNTS")

tab <- tabulator(
  dat,
  rows = c("AEBODSYS", "AEDECOD"),
  columns = "ARM",
  `n` = as_paragraph(fmt_n_percent(n_usub, pct)),
  `nevents` = as_paragraph(n_events)
)

.ARM_COUNTS <- ARM_COUNTS
```

flextable creation:

```{r}
ft <- as_flextable(tab, spread_first_col = TRUE, ) |> 
  prepend_chunks(
    i = ~ is.na(AEBODSYS), j = 1, 
    as_chunk("\t")
  )

for(ARM_COD in names(ARM_COUNTS)){
  ft <- append_chunks(
    x = ft, 
    part = "header", 
    i = 1, 
    j = tabulator_colnames(tab, columns = "n", ARM %in% !!ARM_COD),
    as_chunk(ARM_COUNTS[ARM_COD], 
             formatter = fmt_header_n)
  )
}
ft <- labelizor(
    ft, 
    j = "AEDECOD", 
    part = "body", 
    labels = AEDECOD_LABELS) |> 
  labelizor(part = "header", labels = HEADER_LABELS) |> 
  align(j = 1, align = "left") |> 
  autofit()

ft
```

#### tables

The `tables` package normally generates tables from single datasets, while this
kind of table requires information from two:  `adsl` and `ex_adae`.
One way to handle this would be to add the `adsl` patient count 
information to a copy of the `ex_adae` table.  In this code we
use a different approach:  we generate one table of patient counts
to produce the heading lines, and a second table with the adverse
event data, then use `rbind` to combine the two tables.

```{r}
resetSession()

library(tables)
table_options(doCSS = TRUE)

ex_adae <- formatters::ex_adae

subject_counts <- table(adsl$ARM)

countpercentid <- function(num, ARM) {
  n <- length(unique(num))
  sprintf("%d (%.2f%%)", 
          length(unique(num)), 
          100*n/subject_counts[ARM[1]])
}

count <- function(x) sprintf("(N=%d)", length(x))

heading <- tabular(Heading("")*1*
                     Heading("")*count ~ 
                   Heading()*ARM, data = adsl)

body <- tabular( Heading("Patients with at least one event")*
                   countpercentid*Arguments(ARM = ARM)*
                   Heading()*USUBJID +
                 Heading("Total number of events")*1 +
                 Heading()*AEBODSYS*
                   (Heading("Patients with at least one event")*
                      Percent(denom = ARM, fn = countpercentid)*
                      Heading()*USUBJID +
                    Heading("Total number of events")*1 +
                    Heading()*AEDECOD*DropEmpty()) ~ 
                 Heading()*ARM, 
                 data = ex_adae )

rbind(heading, body)
```
