--- 
title: "Tables in Clinical Trials with R"
author: "R Consortium Tables Working Group"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography:
- book.bib
- packages.bib
description: |
  Example code to create tables commonly used for analyzing clinical trials data.
link-citations: yes
github-repo: "RConsortium/rtrs-wg"
editor_options:
  chunk_output_type: console
---

# About

## Introduction

In this book we present various aspects of creating tables with the `R` language
to analyze and report clinical trials data. The book is initiated by the
R Consortium working group *R Tables for Regulatory Submissions (RTRS)*.

The R Consortium working group *R Tables for Regulatory Submissions (RTRS)* includes
representation from several large pharmaceutical companies and CRO’s. The goal
of the working group is to create standards for creating tables that meet the
requirements of FDA submission documents, and hence enhance the suitability of R
for FDA submissions. It is part of a larger R Consortium effort to facilitate
the certification and validation of R packages and tools for FDA submissions
thereby allowing drug developers to submit documentation for regulatory approval
using the R programming environment in conjunction with open-source packages
without the need for closed and often expensive proprietary tools.


## Call for Contributions

The content of this book is intended to grow via community contribution, so please
add your subject matter expertise to this content by cloning the [github repository of this book](https://github.com/RConsortium/rtrs-wg) and make a pull request with your changes.

We welcome all contribution, including but not limited to:

 - summarizing table packages 
 - adding new example tables
 - clarifying requirements and analyses
 - improving R code

In case you are new to using git and GitHub but would like to make a
contribution then please write a GitHub issue and we will reach out to you to
add the content.


The most convenient way to get started is to:

  - clone the [GitHub repository of this book](https://github.com/RConsortium/rtrs-wg)
  - install the [RStudio IDE](https://www.rstudio.com/products/rstudio/download/)
  - Open the RStudio project in `rtrs-wg/tables-book/tables-book.Rproj`
  - start working, see the [Bookdown documentation](https://bookdown.org/yihui/bookdown/), you can develop the book with the `bookdown::serve_book()` call.
  - once ready to share your work [create a PR](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/proposing-changes-to-your-work-with-pull-requests/creating-a-pull-request)
  - use [GitHub issues](https://github.com/RConsortium/rtrs-wg/issues) to communicate with us if you have difficulties


## Data Used For Examples

We use synthetic data for the examples in this book. The data is available from the
`random.cidsc.data` R package which contains a number of datasets that follow the 
[CDISC ADaM specifications](https://www.cdisc.org/standards/foundational/adam).

```{r, eval=FALSE}
remotes::install_github("insightsengineering/random.cdisc.data")
```

The data in `random.cdisc.data` is completely synthetic, meaning no patient data
has been used to create it. The data is also fairly basic, meaning real study 
data often has more signal and patterns.

```{r}
data(package="random.cdisc.data")$results[, "Item"]
```

In this document, the prepending `c` stands for caches. So, for example the cached 
synthetic subject level dataset `ADSL`:

```{r}
data("cadsl", package = "random.cdisc.data")
head(cadsl)

adsl <- cadsl
```



## Installing the R packages

At some point we may switch to `renv` to install the `R` packages used 
for this book. For right now you can install the packages yourself with:

```{r, eval = FALSE}
install.packages(c("rtables", "remotes", "tidyverse", "bookdown", "r2rtf"))
install.packages('flextable', repos = 'https://davidgohel.r-universe.dev')

remotes::install_github("rstudio/gt")
remotes::install_github("insightsengineering/random.cdisc.data")
remotes::install_github("insightsengineering/scda")
remotes::install_github("insightsengineering/nestcolor")
remotes::install_github("insightsengineering/tern", ref = "v0.7.10")
remotes::install_github("dmurdoch/tables")
```

In each of the sections below, we will reset R to close to the
present state
at the start of the section, so readers can execute the demonstration
code more or less independently of the other sections.  This is done using
the functions defined below.  In your own documents, you wouldn't 
need these resets.

```{r include=FALSE}
# Get the list of attached packages and global environment objects
.initial_state <- currentState()
```

